---
title: "Marginal likelihood estimation for NLME models in dMod: EM, FOCEI, and adaptive sparse-grid Gauss-Hermite quadrature"
author: "Simon Beyer"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::pdf_document:
    toc: false
    number_sections: true
    keep_tex: false
header-includes:
  - \usepackage{amsmath,amssymb}
  - \usepackage{bm}
  - \usepackage{booktabs}
vignette: >
  %\VignetteIndexEntry{Marginal likelihood estimation for NLME models in dMod}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE,
  echo     = TRUE
)
```

\setcounter{tocdepth}{3}
\tableofcontents
\newpage

# Introduction

Nonlinear mixed-effects (NLME) models are the standard tool of population
pharmacokinetics, pharmacodynamics, and other settings where each subject
contributes a short, noisy trajectory of an underlying nonlinear
dynamical system. The challenge of estimating their parameters is
fundamentally the challenge of computing or approximating an intractable
integral: the marginal likelihood.

This vignette has three parts. The first introduces the NLME model in
the ODE modelling context that dMod is built for, and writes down the
marginal likelihood. The second develops the Expectation-Maximisation
(EM) algorithm rigorously, since both estimators in dMod are
particular instances of it. The third covers the two estimators
themselves:

* **FOCEI** (first-order conditional estimation with interaction), in
  which the E-step is replaced by a Laplace (one-point) approximation
  at the per-subject posterior mode.
* **Adaptive sparse-grid Gauss-Hermite quadrature**, in which the
  E-step is replaced by a multi-point Smolyak rule adapted to the
  per-subject posterior, embedded in an ECM (expectation conditional
  maximisation) outer loop.

The code references in this vignette point at specific entries in the
package source so that the implementation can be audited against the
math.

# The NLME model in the dMod / ODE context

dMod composes a fit problem from four named function-objects:

* `x(times, pars)` solves the ODE
  $$
  \dot{\mathbf{x}}(t) \;=\; f\!\left(\mathbf{x}(t),\, \phi\right),
  \qquad \mathbf{x}(0) \;=\; \mathbf{x}_0(\phi),
  $$
  with state vector $\mathbf{x}$ and parameter-dependent right-hand
  side $f$. Code-generated and compiled to native code through
  `odemodel()`. Hereafter $f$ is reserved for this RHS.
* `g(...)` is the *observation function* mapping the ODE solution to
  observables,
  $$
  g\!\left(\mathbf{x}(t; \phi)\right) \;\in\; \mathbb{R}^{d_y}.
  $$
  Concentrations, transformed states, ratios, sums; whatever the
  experiment actually measures. Hereafter $g$ is reserved for this
  observation map.
* `p(...)` is the *parameter transformation* mapping outer parameters
  to inner ODE/observation parameters. For NLME this is exactly the
  place where the population structure
  $\phi_i = h(\theta, \eta_i)$ enters: typically
  $\phi_{i,k} = \exp(\theta_k + \eta_{i,k})$ for log-normal random
  effects, with `branch(trafo, table = subj_table, apply = "insert")`
  routing per-subject etas (see `R/parameters.R`).
* `errmodel` defines the residual variance structure, with
  $\sigma(\phi_i, t_{ij})$ optionally depending on parameters and time.

The composed prediction function $g \circ x \circ p$ produces, for
subject $i$ at time $t_{ij}$,

$$
\mu_{ij}(\psi, \eta_i) \;=\; g\!\left(\mathbf{x}\!\left(t_{ij};\, h(\theta, \eta_i)\right)\right),
$$

and the observation model is

\begin{equation}
y_{ij} \;=\; \mu_{ij}(\psi, \eta_i) \;+\; \varepsilon_{ij},
\qquad \varepsilon_{ij} \;\sim\; \mathcal{N}\!\left(0,\, \sigma^2(\phi_i, t_{ij})\right),
\qquad \eta_i \;\sim\; \mathcal{N}\!\left(0,\, \Omega\right).
\label{eq:nlme}
\end{equation}

The population parameters are $\psi = (\theta, \Omega, \sigma)$. The
NLME-specific parts of \eqref{eq:nlme}, the random effects $\eta_i$
and the covariance $\Omega$, sit *outside* the ODE: $f$ does not see
$\eta_i$ at all; $\eta_i$ only enters through
$\phi_i = h(\theta, \eta_i)$, which is what `p()` evaluates. This
separation matters for two reasons:

1. Sensitivities $\partial \mu_{ij} / \partial \eta_{i,k}$ come from
   the standard dMod sensitivity pipeline composed via the chain rule
   $\partial \mu / \partial \eta = \partial g / \partial x \cdot \partial x / \partial \phi \cdot \partial \phi / \partial \eta$.
   No separate machinery is needed for the random effects.
2. The marginal likelihood (next section) factorises over subjects
   because each $\eta_i$ couples only to $y_i$ via the ODE and back to
   the population through the prior $\mathcal{N}(0, \Omega)$.

The marginal log-likelihood is, summing over subjects,

\begin{equation}
\ell(\psi) \;=\; \sum_{i=1}^N \log L_i(\psi),
\qquad
L_i(\psi) \;=\; \int p\!\left(y_i \mid \eta_i, \psi\right) \, p\!\left(\eta_i \mid \Omega\right) \, d\eta_i.
\label{eq:marginal}
\end{equation}

For an ODE-based $\mu_{ij}$ this integral has no closed form. Every
NLME estimator is a particular strategy for approximating
\eqref{eq:marginal} cheaply enough to be evaluated thousands of times
during outer optimisation while still being faithful enough to recover
the true maximum.

In code, the integrand of \eqref{eq:marginal} is built by composing
`normL2(data, g * x * p, errmodel = err)`, which provides
$-2 \log p(y_i \mid \eta_i, \psi)$ summed across subjects, with
`constraintL2(mu = 0, Omega = om)`, which provides the MVN prior
$-2 \log p(\eta_i \mid \Omega)$ including the $N \log |\Omega|$
normalisation. Their sum is the unintegrated joint negative
log-density $-2 \log p(y, \eta \mid \psi)$ that both estimators
consume. The exact call is shown in the example at the end of this
vignette.

# The Expectation-Maximisation algorithm

The marginal likelihood \eqref{eq:marginal} couples observed data $y$
and latent variables $\eta$. EM is the canonical algorithm for
maximising likelihoods of this form. Both FOCEI and the quadrature
method are particular approximations of it; we therefore develop EM
in full first, then specialise.

## Jensen's lower bound

Pick any reference parameter value $\psi^{(t)}$ and let
$q_i(\eta_i) \;=\; p(\eta_i \mid y_i, \psi^{(t)})$ be the posterior of
$\eta_i$ given the data under $\psi^{(t)}$. For any other $\psi$,

\begin{align*}
\log L_i(\psi)
\;&=\; \log \int q_i(\eta_i) \, \frac{p(y_i, \eta_i \mid \psi)}{q_i(\eta_i)} \, d\eta_i \\
\;&\geq\; \int q_i(\eta_i) \, \log \frac{p(y_i, \eta_i \mid \psi)}{q_i(\eta_i)} \, d\eta_i
&&\text{(Jensen)} \\
\;&=\; \underbrace{\mathbb{E}_{q_i}\!\left[\log p(y_i, \eta_i \mid \psi)\right]}_{Q_i(\psi \mid \psi^{(t)})}
\;-\; \underbrace{\mathbb{E}_{q_i}\!\left[\log q_i(\eta_i)\right]}_{\text{entropy, independent of }\psi}.
\end{align*}

The inequality is *tight* when $\psi = \psi^{(t)}$, because then
$p(y_i, \eta_i \mid \psi) / q_i(\eta_i) = p(y_i \mid \psi^{(t)})$,
which is constant in $\eta_i$ and Jensen becomes an equality.

Summing across subjects, the population $Q$-function is

$$
Q(\psi \mid \psi^{(t)}) \;=\; \sum_{i=1}^N Q_i(\psi \mid \psi^{(t)})
\;=\; \sum_{i=1}^N \mathbb{E}_{q_i}\!\left[\log p(y_i, \eta_i \mid \psi)\right].
$$

## The EM iteration

EM alternates between

* **E-step:** at $\psi^{(t)}$, evaluate $q_i = p(\eta_i \mid y_i, \psi^{(t)})$
  and assemble $Q(\psi \mid \psi^{(t)})$ as a function of $\psi$.
* **M-step:** $\psi^{(t+1)} = \arg\max_\psi Q(\psi \mid \psi^{(t)})$.

## Monotone improvement

The EM iterate never decreases the marginal log-likelihood. From the
Jensen bound,

$$
\ell(\psi^{(t+1)}) \;\geq\; Q(\psi^{(t+1)} \mid \psi^{(t)}) - \mathcal{H}(q)
\;\geq\; Q(\psi^{(t)} \mid \psi^{(t)}) - \mathcal{H}(q)
\;=\; \ell(\psi^{(t)}),
$$

where the first inequality is Jensen at $\psi^{(t+1)}$, the second
follows from the M-step (since $\psi^{(t+1)}$ maximises $Q$, and
$\mathcal{H}(q) = -\mathbb{E}_q \log q$ does not depend on $\psi$),
and the equality is tightness at $\psi = \psi^{(t)}$. Strict
improvement unless the current iterate is already a stationary point
of $\ell$.

## Expectation Conditional Maximisation (ECM)

When the M-step is itself a hard nonlinear problem, but $\psi$ splits
into blocks $\psi = (\psi_A, \psi_B)$ such that the conditional
updates $\psi_A \mid \psi_B$ and $\psi_B \mid \psi_A$ are easy,
replace the M-step with a sequence of conditional maximisations:

* **CM-1:** $\psi_A^{(t+1)} \;=\; \arg\max_{\psi_A} \; Q\!\left((\psi_A, \psi_B^{(t)}) \mid \psi^{(t)}\right)$.
* **CM-2:** $\psi_B^{(t+1)} \;=\; \arg\max_{\psi_B} \; Q\!\left((\psi_A^{(t+1)}, \psi_B) \mid \psi^{(t)}\right)$.

Each CM step still increases $Q$, hence each ECM cycle still increases
$\ell$. The blockwise structure is critical for NLME models where the
variance component $\Omega$ admits a closed-form conditional update
from sufficient statistics while the structural parameters $\theta$
do not.

## Specialising to NLME

In our setting
$\log p(y_i, \eta_i \mid \psi) = \log p(y_i \mid \eta_i, \psi) + \log p(\eta_i \mid \Omega)$,
both pieces depend on $\psi$, and the latent posterior is the
nonstandard

$$
q_i(\eta_i) \;\propto\; p(y_i \mid \eta_i, \psi^{(t)}) \, p(\eta_i \mid \Omega^{(t)}).
$$

Two things make this difficult:

1. The E-step expectation
   $\mathbb{E}_{q_i}[\log p(y_i, \eta_i \mid \psi)]$ is, as a function
   of $\psi$, an integral against a posterior that itself has no closed
   form because $p(y_i \mid \eta_i, \psi)$ runs through the nonlinear
   ODE solution.
2. The M-step still needs derivatives of $Q$ to drive a population
   solver.

Every NLME estimator is a recipe for replacing the intractable E-step
expectation with something computable. Two natural categories of
deterministic methods:

* **Low-order:** approximate $q_i$ by a Gaussian centred at its mode
  and replace the expectation by a one-point rule at that mode. This
  is FOCEI.
* **High-order:** approximate the expectation by a multi-point
  quadrature rule adapted to the local Gaussian approximation of
  $q_i$. This is the sparse-grid Gauss-Hermite path.

Stochastic alternatives (SAEM, etc.) replace the expectation by a
Monte Carlo average drawn from $q_i$ via MCMC. dMod does not
currently implement a stochastic path.

# FOCEI

## Laplace approximation in the E-step

Fix $\psi$ and define the per-subject negative joint log-density

$$
\Phi_i(\eta) \;\equiv\; -\log p\!\left(y_i \mid \eta, \psi\right) -\log p\!\left(\eta \mid \Omega\right).
$$

$\Phi_i$ is smooth and locally convex: the prior pulls it towards the
origin, the likelihood towards the empirical-Bayes optimum. Let
$\hat\eta_i = \arg\min_\eta \Phi_i(\eta)$ and
$H_i = \nabla^2_\eta \Phi_i(\hat\eta_i)$. Expanding $\Phi_i$ to second
order around $\hat\eta_i$ and integrating the resulting Gaussian gives
the Laplace approximation,

\begin{equation}
L_i(\psi) \;\approx\; (2\pi)^{K/2} \, |H_i|^{-1/2} \, \exp\!\left(-\Phi_i(\hat\eta_i)\right).
\label{eq:laplace}
\end{equation}

Taking $-2 \log$ gives the per-subject objective function value (OFV)

$$
\mathrm{OFV}_i(\psi) \;\approx\; 2\, \Phi_i(\hat\eta_i) \;+\; \log |H_i| \;-\; K \log(2\pi).
$$

Summed across subjects, this is the surrogate dMod minimises in place
of $-2 \ell(\psi)$. The constant drops out under optimisation.

The "I" in FOCEI is the *interaction* between the random effects and
the residual variance: $H_i$ is the *full* Hessian of $\Phi_i$ at
$\hat\eta_i$, which couples structural and error-model contributions
through the residual second-derivative term. Methods that drop this
coupling (FOCE without I) lose accuracy whenever $\sigma$ depends on
$\eta$.

## Why this is approximate EM

In the EM language of section 4, FOCEI uses a one-point rule for the
E-step expectation. Specifically, $q_i$ is approximated by the local
Gaussian
$\tilde q_i(\eta_i) = \mathcal{N}(\eta_i \mid \hat\eta_i, H_i^{-1})$
and the expectation is replaced by a single evaluation at
$\hat\eta_i$ plus the normalising factor $\log |H_i|$ from
\eqref{eq:laplace}. The "M-step" then becomes a single combined
optimisation over $\psi$, since the surrogate is now an analytic
function and there is no benefit from conditional splitting.

| EM ingredient | FOCEI approximation |
|---|---|
| posterior $q_i(\eta_i) = p(\eta_i \mid y_i, \psi)$ | Gaussian centred at $\hat\eta_i$ with covariance $H_i^{-1}$ |
| E-step expectation | Laplace one-point rule at $\hat\eta_i$ with $\log |H_i|$ correction |
| M-step | gradient-based outer optimisation of $\sum_i \mathrm{OFV}_i$ |

## Bilevel optimisation

In code, FOCEI is a bilevel problem. The R-side reference
implementation is `R/nlme.R::.emObjfn_laplace`. Per call to the outer
objective at $\psi$:

1. **Inner loop** (one trust-region solve per subject) finds
   $\hat\eta_i(\psi) = \arg\min_\eta \Phi_i(\eta; \psi)$ using dMod's
   exact analytical gradient and Hessian, warmstarted from a
   per-subject cache populated by the previous outer iteration. After
   convergence an eigen-floor is applied,
   $$
   \lambda_k(H_i) \;\leftarrow\; \max\!\left(\lambda_k(H_i),\; \varepsilon\right),
   \qquad \varepsilon \;=\; 10^{-10} \frac{\operatorname{tr} H_i}{K},
   $$
   to guarantee a well-defined $\log |H_i|$ and an invertible Hessian
   at parameter values where the local curvature is degenerate.

2. **Population objective** assembles
   $$
   F(\psi) \;=\; \sum_{i=1}^N \left[ 2\, \Phi_i\!\left(\hat\eta_i\right) \;+\; \log |H_i| \right].
   $$

3. **Outer gradient** is computed via the envelope theorem. Because
   $\hat\eta_i$ minimises $\Phi_i$ in $\eta$, the implicit dependence
   $\hat\eta_i(\psi)$ contributes zero to
   $\nabla_\psi \Phi_i(\hat\eta_i)$, so the only $\eta$-derivative
   that survives is the explicit one,
   $$
   \nabla_\psi F \;=\; 2 \sum_i \nabla_\psi \Phi_i\!\big|_{\eta=\hat\eta_i} \;+\; \sum_i \frac{d \log |H_i|}{d \psi}.
   $$
   The second term is the *Stage-2 correction* (section 5.4).

4. **Outer Hessian** uses the Schur complement of the block-Hessian
   of $\sum_i \Phi_i(\hat\eta_i)$,
   $$
   \nabla^2_\psi F \;\approx\; 2 \sum_i \left[ H_i^{\psi\psi} - H_i^{\psi\eta} \left(H_i^{\eta\eta}\right)^{-1} \left(H_i^{\psi\eta}\right)^{\!\top} \right],
   $$
   where the blocks come from a single batched `joint(deriv = TRUE)`
   call evaluated at every subject's mode.

5. **Outer trust region** (dMod's `trust()`) takes a Newton step on
   $F(\psi)$, the inner cache is updated with the new $\hat\eta_i$
   values, and the loop repeats.

## The Stage-2 analytical correction

dMod's Hessian in step 4 is built from the Gauss-Newton form
$H_{i,\mathrm{GN}} = 2 \nabla_\eta r_i^\top \Sigma_i^{-1} \nabla_\eta r_i + 2 \Omega^{-1}$,
not the true Newton Hessian. Under the GN form the surrogate
$\log |H_i|$ is still needed in the objective and its gradient
$d \log |H_i| / d\psi$ is no longer immediate from the inner solve,
because the GN approximation introduces a discrepancy that vanishes
only at a perfect fit.

`R/nlme.R::.computeFoceiCorrection` computes the closed-form
$d \log |H_{i,\mathrm{GN}}| / d\psi$ via

$$
\frac{d \log |H_i|}{d \psi_k}
\;=\;
\operatorname{tr}\!\left(H_i^{-1} \frac{\partial H_i}{\partial \psi_k}\bigg|_\eta\right)
\;+\;
\sum_l \operatorname{tr}\!\left(H_i^{-1} \frac{\partial H_i}{\partial \eta_l}\right) \cdot \frac{d \hat\eta_l}{d \psi_k}.
$$

The chain term $d \hat\eta / d \psi$ comes from the implicit function
theorem applied to the inner first-order condition: the exact
(Newton) Hessian closes the IFT block, while $H_{i,\mathrm{GN}}^{-1}$
is used in the traces. The required mixed third derivatives are
provided by dMod's deriv2 pipeline (`_s2` files emitted by `CppODE`
and `cOde`).

The correction is recomputed at every outer call so the gradient
remains exactly the gradient of $F(\psi)$. This matches NONMEM /
Monolix / `nlmixr2` semantics, and is what cross-tool comparisons
expect.

## The C++ FOCEI kernel

The bilevel orchestration runs entirely in C++
(`src/focei_kernel.cpp`). The inner per-subject trust region inlines
the residual, the Fisher-information Hessian (Gauss-Newton on the
mean plus the $\partial \log \sigma / \partial \eta$ contribution),
and the MVN $\eta$-prior directly in C++, bypassing
`normL2` / `constraintL2` / `evalConditionResidual` per call. Model
and error functions are still invoked via R (they wrap compiled
native code anyway), but the per-iteration glue is gone. Multi-output
models and $\eta$-dependent $\sigma$ (proportional or combined error
models) are handled in the same loop: the kernel reads
$\partial \sigma / \partial \eta$ from the error model's `deriv`
attribute and propagates it through the per-row gradient and
Hessian.

# Adaptive sparse-grid Gauss-Hermite quadrature

## Why move beyond Laplace

The Laplace approximation \eqref{eq:laplace} treats $\Phi_i$ as
locally quadratic. Its error is $O(K / n_i)$ per subject and is
dominated by the third and fourth derivatives of $\Phi_i$ at the
mode. Two practical scenarios where this matters:

* Datasets with informative-but-short trajectories (oral PK at the
  absorption limit, sparse sampling near saturation) where $\Phi_i$
  is noticeably skewed.
* Variance components $\Omega$ that are themselves the target of
  inference. Laplace systematically over-estimates the marginal
  likelihood because $|H_i|^{-1/2}$ overweights the peak relative to
  the tails, which biases $\Omega$ downward.

A multi-point quadrature rule integrating $L_i$ exactly when its
log-density is a polynomial of bounded degree fixes both, at the cost
of an exponential factor in $K$ if you use a full tensor-product
grid. The Smolyak construction sidesteps that factor.

## Smolyak sparse-grid Gauss-Hermite

For fixed dimension $K$ and level $L \geq K$, the Smolyak rule
combines tensor products of 1D physicists' Gauss-Hermite rules,

$$
A(L, K) \;=\; \sum_{\substack{|\mathbf{i}| = q \\ L-K+1 \leq q \leq L}} (-1)^{L-q} \binom{K-1}{L-q} \; \mathcal{U}^{i_1} \otimes \mathcal{U}^{i_2} \otimes \cdots \otimes \mathcal{U}^{i_K},
$$

where $\mathcal{U}^l$ is the 1D level-$l$ Gauss-Hermite rule with
$m(l) = 2l - 1$ nodes and weight $\exp(-z^2)$. 1D nodes and weights
come from Golub-Welsch: eigendecompose the symmetric tridiagonal
Hermite Jacobi matrix (diagonal $0$, off-diagonal $\sqrt{i/2}$) using
LAPACK `dstev`. The node count grows as $O(K^{L - K})$ rather than
the full-tensor $O(m(L)^K)$.

`src/quadrature.cpp::sparse_grid_gh(K, level)` implements this. It
returns a batch-first $[B, K]$ matrix of $z$-space nodes and a
length-$B$ vector of signed weights. Signs appear because
Gauss-Hermite is non-nested: the Smolyak combination of tensor
products at different levels does not fully cancel at coincident
nodes. The downstream evaluator handles mixed-sign aggregation via
signed log-sum-exp; see section 6.4.

Typical node counts for $K = 3$:

\begin{center}
\begin{tabular}{rrl}
\toprule
Level & Nodes & Comment \\
\midrule
3 & 1   & single-point, trivial \\
4 & 7   & first useful level \\
5 & 31  & polynomial exactness degree $\sim$ 9 \\
6 & 105 & polynomial exactness degree $\sim$ 11 \\
\bottomrule
\end{tabular}
\end{center}

## Adapting to the local posterior

The raw Smolyak rule integrates against the weight
$\exp(-\mathbf{z}^\top \mathbf{z})$ on $\mathbb{R}^K$. To target the
subject-specific posterior $p(\eta_i \mid y_i, \psi)$ we recentre and
rescale around the local Gaussian approximation
$\mathcal{N}(\hat\eta_i, H_i^{-1})$.

Let $L_H$ be the lower Cholesky factor of $H_i$. The change of
variable

$$
\eta \;=\; \hat\eta_i + \sqrt{2} \, L_H^{-\top} \mathbf{z}
$$

maps $z$-space nodes onto $\eta$-space nodes and folds the Jacobian
$2^{K/2} / |\det L_H|$ plus the implicit $\exp(\mathbf{z}^\top \mathbf{z})$
back-correction into augmented weights,

$$
\log |W_b| \;=\; \tfrac{K}{2} \log 2 \;-\; \log |\det L_H| \;+\; \log |w_b^{\mathrm{GH}}| \;+\; \mathbf{z}_b^\top \mathbf{z}_b,
$$

with sign tracked separately. `R/quadrature.R::makeSubjectNodes`
performs this transform. The result is a per-subject grid that puts
most of its mass where the joint posterior actually lives, recovers
exact answers for products of Gaussians with the local approximation
as a special case, and reduces gracefully to Laplace at level $K$ (a
single node at the mode).

## Signed log-sum-exp aggregation

For numerical stability the per-subject log-likelihood is built in
log space. For nodes $b = 1, \dots, B$ define

$$
\ell_b \;=\; \log |W_b| \;-\; \tfrac{1}{2} \Phi_i(\eta_b),
\qquad
s_b \;=\; \operatorname{sgn}(W_b).
$$

Then

$$
\log L_i \;=\; \log \!\left| \sum_b s_b \exp \ell_b \right|
\;=\; M_i \;+\; \log \!\left| \sum_b s_b \exp(\ell_b - M_i) \right|,
\qquad M_i \;=\; \max_b \ell_b.
$$

Downstream moments use the same softmax weights with carried sign,

$$
w^{\mathrm{sm}}_b \;=\; s_b \exp(\ell_b - \log L_i),
\qquad
\hat m_i \;=\; \sum_b w^{\mathrm{sm}}_b \, \eta_b,
\qquad
\hat M_i \;=\; \sum_b w^{\mathrm{sm}}_b \, \eta_b \eta_b^\top.
$$

If $\sum_b s_b \exp(\ell_b - M_i)$ is non-positive or non-finite, the
grid is under-resolved: the evaluator warns and asks for a higher
level. The effective sample size
$1 / \sum_b (w^{\mathrm{sm}}_b)^2$ is exported in the per-subject
diagnostics so grid adequacy can be monitored per subject.

`R/nlme.R::ecmEvaluateSubject` implements the inner kernel, including
the gradient and Hessian pieces below.

## Gradient and Hessian through the integral

The marginal log-likelihood gradient is, at node level,

$$
\nabla_\psi \log L_i \;=\; \sum_b w^{\mathrm{sm}}_b \, \nabla_\psi \ell_b,
\qquad
\nabla_\psi \ell_b \;=\; -\tfrac{1}{2} \nabla_\psi \Phi_i(\eta_b),
$$

a softmax-weighted average. The Hessian uses the log-sum-exp identity

$$
\nabla^2_\psi \log L_i \;=\; \underbrace{\sum_b w^{\mathrm{sm}}_b \, \nabla^2_\psi \ell_b}_{\text{average Hessian}} \;+\; \underbrace{\operatorname{Cov}_{w^{\mathrm{sm}}} (\nabla_\psi \ell_b)}_{\text{spread term}},
$$

where the covariance is taken with $w^{\mathrm{sm}}$ as a signed
probability measure,
$\operatorname{Cov}(g) = \mathbb{E}[g g^\top] - \mathbb{E}[g]\mathbb{E}[g]^\top$.
The identity is algebraically exact for signed weights (it uses only
that the weights sum to one).

Per-node $\nabla_\psi \ell_b$ and $\nabla^2_\psi \ell_b$ come from a
single call to the prediction function plus `evalConditionResidual`.
dMod's existing analytical sensitivity chain through the ODE solver
supplies the derivatives, so the quadrature kernel itself emits pure
numbers and does not need to know anything about the model.

## ECM orchestration

Combining a non-trivial integration rule with population-level
optimisation needs more care than FOCEI. Joint optimisation over
$\psi$ with the modes recomputed every inner step would either
re-locate the grid constantly (expensive) or keep it frozen and stall
on the wrong objective (biased). The classical fix is the ECM
decomposition of section 4.4: per outer pass, split $\psi$ into
blocks and update each block conditionally on the others held fixed.

`R/nlme.R::.runQuadratureEcm` runs the following per stage:

1. **E-step** (refresh quadrature, `R/nlme.R::.emObjfn_quadrature::rebuildQuadrature`).
   Fix the current $\psi$. For each subject, run an inner trust solve
   to find $\hat\eta_i(\psi)$ and the eigen-floored $H_i$. Build the
   adapted Smolyak grid. Freeze the grid for the rest of the stage.
2. **CM-1** (structural). With the grid frozen, run an outer trust on
   the structural parameters $\theta$ (and $\sigma$), using the
   gradient and Hessian from section 6.5.
3. **CM-2** (omega). With $\theta$ frozen at the CM-1 minimiser,
   update $\Omega$ in closed form via the empirical second moments
   $\{\hat M_i\}$.
4. Iterate CM-1 / CM-2 within the stage until $|\Delta \psi|$ or
   $|\Delta \mathrm{OFV} / \mathrm{OFV}|$ falls below tolerance.

After convergence, the stage advances to a higher quadrature level,
re-runs the E-step at the new $\psi$, and repeats CM-1 / CM-2. By
default the stage levels are $\{K{+}1, K{+}2, K{+}3\}$, which for
$K = 3$ gives $\{4, 5, 6\}$ corresponding to $\{7, 31, 105\}$ nodes
per subject.

The default outer mode is `foceiQuadrature`: run FOCEI to convergence
first, then use the converged $\psi$ and $\hat\eta_i$ as warmstart
for the quadrature ECM, so the first E-step starts near the right
grid centres. The interpretation is that FOCEI provides a fast,
low-accuracy E-step (Laplace) used to bootstrap; the subsequent
quadrature ECM provides a higher-accuracy E-step used to refine the
estimate.

## Closed-form omega update

The full-data log-density of the random effects is

$$
\log p(\eta \mid \Omega) \;=\; -\tfrac{1}{2} \sum_i \left[ \log |\Omega| \;+\; \eta_i^\top \Omega^{-1} \eta_i \right] \;+\; \text{const}.
$$

Taking the expectation under the current grid replaces
$\eta_i \eta_i^\top$ by its posterior expectation $\hat M_i$. Setting
the derivative with respect to $\Omega$ to zero gives the sufficient
statistic

$$
S \;=\; \frac{1}{N} \sum_{i=1}^N \hat M_i.
$$

`R/omegaClass.R::updateOmegaChol` consumes $S$ and re-parameterises
into the Cholesky entries that `omegaSpec` exposes. Three structural
cases:

* **Pure diagonal** (`structure = "diag"`):
  $\omega_{kk} = \tfrac{1}{2} \log S_{kk}$, the log-Cholesky entry
  that satisfies $L_{kk}^2 = S_{kk}$.
* **Full unconstrained**: take the Cholesky factor of $S$,
  $L = \mathrm{chol}(S)^\top$, and read entries off positionally;
  log-Cholesky on the diagonal, raw entry off-diagonal.
* **Selective correlate / structural constraints**: no closed form.
  Fall back to a small inner trust on the Q-function,
  $$
  \tfrac{1}{2}\, N \!\left(\log |\Omega| \;+\; \operatorname{tr}(\Omega^{-1} S)\right),
  $$
  with closed-form gradient in $L$; convergence is typically a
  handful of inner iterations.

In all cases CM-2 is a global update (not local around the previous
$\Omega$); it is the exact M-step under the E-step's softmax-weighted
empirical second moment. The deterministic-quadrature analogue of
NONMEM's $\Omega$ update under SAEM, without sampling noise.

## Diagnostics

Each call to the quadrature objective carries an `emDiag` attribute
with components

* `etaModes`, `HiList`: frozen modes and Hessians used to build the
  grid.
* `mHatList`, `MHatList`: first and second posterior moments per
  subject.
* `maxSoftmax`: $\max_b |w^{\mathrm{sm}}_b|$ per subject. A value
  near 1.0 means the integral is dominated by a single node, which
  is fine if the local Gaussian is a good approximation but a warning
  sign otherwise.
* `n_eff`: effective sample size per subject. Falls when the softmax
  concentrates; the ECM loop's `nEffMin` diagnostic surfaces this.
* `nFloored`: per-subject count of eigenvalues clipped at the
  Hessian floor. Persistent nonzero values indicate near-singular
  per-subject Fisher information, usually a missing identifiability
  constraint upstream.
* `level`, `innerConverged`, `innerIter`: bookkeeping.

`plotTrace(fit)` shows the per-stage ECM convergence (OFV vs. ECM
iteration with stage markers). `plotHistIndivs(fit)` overlays
per-subject $\hat\eta_i$ histograms with the QQ line against the
fitted $\mathcal{N}(0, \Omega_{kk})$ and is informative for spotting
mis-specified variance structures.

# End-user API

This chapter collects the relevant function calls. The full
walkthrough on the Theophylline dataset (12 subjects, 1-compartment
oral PK) is `inst/examples/theophylline.R`.

## Building the objective

The joint negative log-density of \eqref{eq:marginal} is built from
the standard dMod composable function pipeline:

```{r joint-objfn}
library(dMod)

# x: ODE solution (from odemodel + Xs).
# g: observation function.
# p: parameter transformation, with branch(..., apply = "insert") to
#    route per-subject etas.
# err: optional error model (errmodel = NULL for fixed sigma in data).
# om: omegaSpec with subject expansion, built via omega(..., subjects = ...).

joint <- normL2(dlist, g * x * p, errmodel = err) +
         constraintL2(mu = 0, Omega = om)
```

The `init` vector must contain every structural parameter, every
log-scale residual SD, and every Cholesky entry `om$cholPars`. See
the Theophylline example for a worked starting vector.

## FOCEI

```{r focei-call}
fit_focei <- nlmeFit(
  joint, om, init,
  model    = model,           # the prdfn g * x * p
  data     = dlist,
  errmodel = err,
  method   = "focei",
  control  = list(focei = list(
    innerControl = list(rtol = 1e-7, maxit = 30),
    trustControl = list(iterlim = 100))))

print(fit_focei)
```

There is one FOCEI path: the C++ kernel with eager Stage-2 correction.
`innerControl` tunes the per-subject inner trust region (warmstart
radius, convergence tolerances, eigenvalue floor); `trustControl`
tunes the outer trust region. Multi-output and $\eta$-dependent
$\sigma$ models are handled by the same loop. There are no other
knobs.

## Quadrature refinement

In the same spirit as a polish step in optimisation, the
`foceiQuadrature` workflow uses FOCEI as a fast low-accuracy
estimator to converge on a parameter neighbourhood, then refines the
estimate with the higher-accuracy quadrature ECM:

```{r quad-call}
fit_quad <- nlmeFit(
  joint, om, init,
  model    = model, data = dlist, errmodel = err,
  method   = "foceiQuadrature",
  control  = list(
    quadrature = list(epsQuadLevels  = c(4, 5, 6),
                      maxEcmPerStage = 5L,
                      maxCm1Iter     = 30L)))

print(fit_quad)
print(fit_quad$stageTrace)
```

`epsQuadLevels` is the sequence of Smolyak depths used by the outer
stages. Defaults to $K{+}1{:}3$ (for $K{=}3$, that is
$\{4, 5, 6\}$, corresponding to $\{7, 31, 105\}$ nodes per subject).
The cost scales linearly in nodes-per-subject and in the number of
E-step refreshes (one per ECM iteration per stage), so problems with
$K > 3$ or levels above $K{+}3$ become expensive quickly. For
exploratory work, FOCEI alone is the recommended starting point;
the quadrature refinement is a confirmatory step once the model
structure is settled.

On the Theophylline example the refinement leaves the structural
means essentially unchanged and corrects the omega-Cholesky entries
downward, the textbook Laplace-bias correction. Quantitative
numbers depend on the exact starting values, the local hardware,
and the BLAS in use.

## Inspecting the fit

`fit_focei` and `fit_quad` are `nlmeFit` S3 objects with the same
field layout (`argument`, `value`, `gradient`, `hessian`, `omega`,
`etaModes`, `converged`, `iterations`, `emDiag`, `method`, ...).
The diagnostic generics are camelCase S3 methods:

```{r diagnostics}
plot          (fit_focei)   # DV vs IPRED + DV vs PRED
plotIndivs    (fit_focei)   # per-subject IPRED/PRED curves
plotResiduals (fit_focei)   # IWRES vs IPRED + IWRES vs TIME
plotHistIndivs(fit_focei)   # eta histogram + QQ vs N(0, Omega_kk)
plotTrace     (fit_quad)    # ECM convergence trace (quadrature only)
```

Per-call quadrature diagnostics (modes, Hessians, moments, effective
node count, softmax mass) live on `attr(em(psi), "emDiag")` and on
`fit_quad$emDiag` after the run.

# References

Wang, Y. (2007). Derivation of various NONMEM estimation methods.
*J. Pharmacokinet. Pharmacodyn.* 34, 575-593.

Pinheiro, J. C., Bates, D. M. (1995). Approximations to the
log-likelihood function in the nonlinear mixed-effects model.
*J. Comput. Graph. Stat.* 4, 12-35.

Smolyak, S. A. (1963). Quadrature and interpolation formulas for
tensor products of certain classes of functions. *Soviet Math. Dokl.*
4, 240-243.

Pinheiro, J. C., Bates, D. M. (2000). *Mixed-Effects Models in S and
S-PLUS*. Springer.

Lavielle, M. (2014). *Mixed Effects Models for the Population
Approach*. Chapman & Hall / CRC.

Meng, X.-L., Rubin, D. B. (1993). Maximum likelihood estimation via
the ECM algorithm: A general framework. *Biometrika* 80, 267-278.

Dempster, A. P., Laird, N. M., Rubin, D. B. (1977). Maximum
likelihood from incomplete data via the EM algorithm.
*J. R. Stat. Soc. B* 39, 1-38.
