---
title: "PEtab problems with dMod"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PEtab problems with dMod}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      dpi = 150, fig.retina = 2, out.width = "100%")
have_libsbml <- file.exists(path.expand("~/.virtualenvs/libsbml/bin/python"))
```

dMod reads [PEtab](https://petab.readthedocs.io/) v1 and v2 problems
(YAML manifest + SBML model + four to six TSV tables) and turns them into
a fully assembled fit problem in one call. This vignette walks through
**Boehm et al. 2014** (JAK/STAT5 signalling, the canonical PEtab
benchmark) using a bundled copy of the files, then explains what
`importPEtab()` does under the hood.

## Prerequisites

`importPEtab()` reads the SBML model through a Python helper that needs
`python-libsbml`. Create the virtualenv once:

```bash
python3 -m venv ~/.virtualenvs/libsbml
~/.virtualenvs/libsbml/bin/pip install python-libsbml
```

The TSV / YAML readers below work without it.

## The bundled example

```{r files}
petab_dir <- system.file("extdata/petab_boehm", package = "dMod")
list.files(petab_dir)
```

```{r yaml}
library(dMod)
yamlPath <- file.path(petab_dir, "Boehm.yaml")
manifest <- read_petab_yaml(yamlPath)
str(manifest, max.level = 2)
```

```{r tables}
tables <- read_petab_tables(yamlPath)
head(tables$measurements)
tables$observables
tables$parameters[, c("parameterId", "parameterScale",
                       "lowerBound", "upperBound", "estimate")]
```

The model has 8 STAT5 species, 6 estimated rate constants, 3 estimated
noise SDs, and 2 fixed scaling parameters (`ratio`, `specC17`). All
estimated parameters are on `log10` scale.

## Importing the problem

`importPEtab()` parses the YAML, the SBML, and all TSV tables, generates
the ODE plus sensitivity equations, compiles them, and returns a single
`PEtabProblem` list:

```{r import, eval = have_libsbml}
setwd(tempdir())  # native artefacts (.c, .so, ...) land in cwd
petab <- importPEtab(yamlPath, solver = "deSolve", cores = 2L)
print(petab)
```

The slots are plain dMod objects:

| slot       | what it is                                                |
|------------|-----------------------------------------------------------|
| `dataList` | measurements as a `datalist`, keyed by sub-condition      |
| `odemodel` | compiled ODE rhs + sensitivities                          |
| `g`, `x`, `p` | observation function, prediction function, trafo       |
| `e`        | error model (Y function over the noise formulas), or `NULL` if every noise formula collapses to a constant after substitution |
| `prd`      | composite `g * x * p`                                     |
| `obj`      | objective `normL2(dataList, prd, errmodel = e)` with PEtab `fixed` parameters baked in |
| `bestfit`  | PEtab `nominalValue` per estimated parameter (the published MLE for benchmark problems, the current best estimate and optimizer starting point otherwise) |
| `parlower`, `parupper` | bounds, on `parameterScale`                   |

## Predict and plot

The prediction `g * x * p` returns one observable trace per timepoint. To
visualise the noise model, we read the per-observable $\sigma$ from
`bestfit` (the $\sigma$ entries are estimated parameters; Boehm declares
`parameterScale = log10`, so we linearise) and overlay $\pm\sigma$ bands
via `geom_ribbon`.

```{r predict, eval = have_libsbml, fig.width = 8, fig.height = 3.5, message = FALSE}
library(ggplot2)

fixed      <- attr(petab, "petab_meta")$fixed
times      <- seq(0, 240, length.out = 121)
prediction <- petab$prd(times, c(petab$bestfit, fixed))

# Linearise per-observable σ from the bestfit's parameterScale.
scales  <- attr(petab$bestfit, "petab_scales")
unscale <- function(p, sc) ifelse(sc == "log10", 10^p,
                           ifelse(sc == "log",   exp(p), p))
sds     <- unscale(petab$bestfit, scales[names(petab$bestfit)])
sd_obs  <- c(pSTAT5A_rel = unname(sds["sd_pSTAT5A_rel"]),
             pSTAT5B_rel = unname(sds["sd_pSTAT5B_rel"]),
             rSTAT5A_rel = unname(sds["sd_rSTAT5A_rel"]))

pred_mat  <- as.data.frame(prediction[[1]])
pred_long <- do.call(rbind, lapply(names(sd_obs), function(o)
  data.frame(name = o, time = pred_mat$time, value = pred_mat[[o]],
             sigma = sd_obs[[o]])))
pred_long$lo <- pred_long$value - pred_long$sigma
pred_long$hi <- pred_long$value + pred_long$sigma

ggplot() +
  geom_ribbon(data = pred_long,
              aes(time, ymin = lo, ymax = hi, fill = name), alpha = 0.2) +
  geom_line(data = pred_long, aes(time, value, color = name)) +
  geom_point(data = petab$dataList[[1]],
             aes(time, value, color = name), size = 1.5) +
  facet_wrap(~ name, scales = "free_y") +
  guides(color = "none", fill = "none") +
  theme_bw() +
  labs(x = "time [min]", y = "observable")
```

## Fit

```{r fit, eval = have_libsbml}
fit <- trust(petab$obj, petab$bestfit, rinit = 1, rmax = 10)
fit$value
```

`obj()` already has the fixed parameters and PEtab parameter scales baked
in, so the optimizer only sees the estimated outer parameters on the
scale declared in `parameters.tsv`.

## How `importPEtab()` works internally

```
   yaml + tsv + sbml
          │
          ▼
   parse / normalise   (v2 → v1 if needed)
          │
          ├── parameters    ──▶  p  =  P(trafo)
          ├── reactions     ──▶  odemodel  ──▶  x  =  Xs(odemodel)
          ├── observables   ──▶  g  =  Y(observables)   +   errmodel e
          └── measurements  ──▶  sub-condition map      +   dataList
                                            │
                                            ▼
                                  prd  =  g * x * p
                                            │
                                            ▼
                          obj  =  normL2(dataList, prd, errmodel = e)
```

Each box maps onto a slot of the returned `PEtabProblem`.

The non-obvious pieces:

**1. SBML extensive → dMod intensive flux.** SBML kinetic laws have
units of amount/time; dMod stores rates as concentration/time. On
import, each kinetic law is divided by the volume of its home
compartment:
$$\text{rate}_{\text{dMod}} \;=\; K_{\text{SBML}} \,/\, V_{\text{home}}.$$
The home compartment is the shared compartment of the educts, falling
back to the products for pure synthesis. AssignmentRules from SBML are
inlined into rates and initial expressions before the eqnlist is built.

**2. `parameterScale` is folded into the trafo, not into the data.**
PEtab declares per-parameter scales (`lin`, `log`, `log10`); dMod wraps
the trafo `p` so the optimizer sees parameters on the chosen scale
($x = 10^{X}$ for `log10`, $x = e^{X}$ for `log`). The chain rule wraps
through the composition automatically, so gradients stay correct. The
wrapping only applies to estimated outer parameters; fixed parameters
stay linear. The `bestfit` vector therefore lives on `parameterScale`,
not on the linear scale.

**3. Per-row observable / noise placeholders.** `measurements.tsv` may
override `observableParameter${k}_${id}` and `noiseParameter${k}_${id}`
placeholders per row (e.g. a separate noise SD per observable, as in
Boehm). Placeholders carry the row's `observableId` as suffix, so
substitutions across observables never collide. Within a
`(simulationConditionId, preequilibrationConditionId)` group the
importer merges all rows into a single sub-condition whose trafo carries
one substitution per observableId. Boehm's three rows therefore collapse
to one sub-condition with three independent noise substitutions. A new
sub-condition is only spawned when the same observableId appears with
conflicting `(observableParameters, noiseParameters)` values inside one
group. Numeric noise placeholders fold into the data column (`data$sigma`,
errmodel stays `NULL`); symbolic ones become inner parameters of the
trafo.

**4. Pre-equilibration.** When measurements carry a
`preequilibrationConditionId`, the importer composes a `Pequil` stage
into the prediction function. Each affected sub-condition runs the
peq-row's parameter overrides through the rates, integrates to steady
state, and feeds the result into the simulation period.

**5. SBML-default-as-fixed gap.** Symbols that appear in rates, inits,
or observables but are not in `parameters.tsv` and are not states are
pulled from SBML's parameter table and appended to `fixed` (typically
compartment volumes that PEtab leaves unspecified). Symbols without a
default raise a clear error.

**6. Error model wiring.** `e` is built once from the raw `noiseFormula`
strings (e.g. Boehm's `noiseParameter1_pSTAT5A_rel`); placeholders
survive verbatim as inner-parameter symbols of the Y function. The trafo
`p` then carries the per-observable substitutions (in Boehm:
`noiseParameter1_pSTAT5A_rel = sd_pSTAT5A_rel`, plus analogues for
`pSTAT5B_rel`, `rSTAT5A_rel`, all in **one** trafo; see point 3). During
`prd = g * x * p` the placeholders bind to the right outer parameters
before `e` reads them, so a single `e` covers every observable in a
merged sub-condition. When every noise formula reduces to a constant
after substitution, `e` is `NULL` and $\sigma$ lives in
`dataList[[…]]$sigma`, taking `normL2`'s fast path without an error
model.

**7. v2 → v1 normalisation.** `.petab_v2_normalize_tables()` adapts v2
manifests to v1 shapes so the downstream trafo / observation / objective
builders are shared. v2 specifics handled here: long-format conditions
pivot to wide; `experiments.tsv` `(time, conditionId)` sequences map to
`(simulationConditionId, preequilibrationConditionId)` (max two periods);
the combined `noiseDistribution` splits into observable transformation +
distribution; explicit `observablePlaceholders` / `noisePlaceholders`
are rewritten to v1's `observableParameter${k}_${id}` sentinels; an
optional `mapping.tsv` is applied as a textual rewrite of PEtab entity
IDs into model entity IDs.

## PEtab v2 feature coverage

The tables below summarise dMod's support relative to the PEtab v2
specification. Items marked ⚠ work with caveats; ✗ items are either
silently ignored or rejected at import.

### SBML model

| feature                                          | dMod | notes                                                                 |
|--------------------------------------------------|:----:|-----------------------------------------------------------------------|
| reactions, kinetic laws (incl. `<power/>`)       | ✓    | extensive `K_SBML` divided by home-compartment volume on import       |
| compartments, multi-compartment                  | ✓    | first-class on `eqnlist`                                              |
| `<initialAssignment>`                            | ✓    | symbolic initials carried verbatim and written on export              |
| `<assignmentRule>`                               | ✓    | inlined into rates / inits / observables before the `eqnlist` is built |
| `<rateRule>` on a species                        | ✓    | appended as a virtual reaction (stoichiometry +1)                     |
| `<rateRule>` on a parameter / compartment        | ⚠    | warned and skipped; promote the symbol to a state                     |
| `<event>` with `time >= T` trigger               | ✓    | rewritten as a dMod event                                             |
| `<event>` with non-time triggers                 | ⚠    | symbolic `root` fallback                                              |
| `<functionDefinition>`                           | ✓    | inlined via libsbml's `SBMLFunctionDefinitionConverter`               |
| `<algebraicRule>`                                | ✗    | silently skipped (dMod has no DAE solver)                             |
| non-SBML model languages (BNGL / CellML / PySB)  | ✗    | only SBML is read                                                     |

### PEtab tables

| feature                                                      | dMod | notes                                                       |
|--------------------------------------------------------------|:----:|-------------------------------------------------------------|
| `parameters.tsv` with `parameterScale`                       | ✓    | `lin` / `log` / `log10`, folded into the trafo              |
| `observables.tsv` with `observableTransformation`            | ✓    | per-observable `lin` / `log` / `log10`                      |
| `conditions.tsv` (long-format)                               | ✓    | pivoted to wide for the trafo builder                       |
| `measurements.tsv` per-row `noiseParameters` / `observableParameters` | ✓ | numeric and symbolic                                       |
| `experiments.tsv` with ≤ 2 periods                           | ✓    | one pre-equilibration plus the simulation period            |
| `experiments.tsv` with > 2 periods                           | ✗    | rejected at import                                          |
| `mapping.tsv`                                                | ✓    | textual rewrite of `petabEntityId` → `modelEntityId`        |
| multi-model (`model_files` with several ids)                 | ✓    | per-model `(odemodel, x, g, p)` composed via `+`; sub-condition keys are prefixed with the model id |
| priors (`priorDistribution`, `priorParameters`)              | ⚠    | only `parameterScaleNormal` and `normal` on `lin`-scale parameters; mapped to `constraintL2` |
| `visualisation.tsv`                                          | ✗    | ignored                                                     |

### Likelihoods and noise

dMod's only objective is `normL2`, so PEtab noise distributions map as:

| `noiseDistribution`         | dMod | notes                                              |
|-----------------------------|:----:|----------------------------------------------------|
| `normal`                    | ✓    | direct                                             |
| `log-normal`                | ✓    | via `observableTransformation = log`               |
| `laplace`, `log-laplace`    | ✗    | rejected (dMod uses L2 only)                       |

### Export caveats

When writing a dMod problem to PEtab v2 via `exportPEtab()` or
`exportPEtabObject()`, a few features cannot be expressed in the v2
schema:

| caveat                                              | notes                                                                |
|-----------------------------------------------------|----------------------------------------------------------------------|
| `parameterScale` on v2 export                       | linearised on disk with a one-time warning (v2 has no `parameterScale` column) |
| `log10` `observableTransformation` on v2 export     | rewritten to `log-normal` noise (v2 has no `log10` observable transformation) |
| state-init overrides from a v1 `conditions.tsv`     | not written to the v2 long-form conditions table; export to `formatVersion = "1"` keeps them |
| `<algebraicRule>` in the source SBML                | absent from the dMod model and therefore from any SBML written back  |

The dMod-native exporter `exportPEtab(data, reactions, observables, p,
pouter, ...)` builds a PEtab problem from raw dMod objects rather than
from an imported one. It supports per-row sigma via the
`noiseParameter1_<obsId>` + `measurements.tsv$noiseParameters` path, but
does not write pre-equilibration through `p`; go via
`exportPEtabObject()` if the source is an imported problem.

### Notes

- **L2 likelihoods only.** PEtab `normal` is handled directly; `log-normal`
  via `observableTransformation = log` (mathematically L2 on the
  log-transformed observable). `laplace` and `log-laplace` are *not*
  supported and won't be: dMod's only objective is `normL2`.
- **Multi-model PEtab v2.** Multiple `model_files` entries with a
  `modelId` column on `measurements.tsv` import as separate per-model
  `(odemodel, x, g, e, p)` pieces, then compose via dMod's `+` so the
  fit problem stays a single objective. Sub-condition keys are prefixed
  with the modelId to keep them disjoint.
- **L2 priors on parameters.** PEtab `priorDistribution` /
  `priorParameters` (v2) and `objectivePriorType` /
  `objectivePriorParameters` (v1) are read into a `constraintL2(mu,
  sigma)` term that is summed into `obj`. Accepted distributions are
  `parameterScaleNormal` (always) and `normal` on `lin`-scale parameters
  (where the two coincide); other distributions raise an error.
