---
title: "Structural identifiability in dMod: symmetries, the observability-identifiability matrix, and the determining equations"
author: "Simon Beyer"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  rmarkdown::pdf_document:
    toc: false
    number_sections: true
    keep_tex: false
    latex_engine: xelatex
header-includes:
  - \usepackage{amsmath,amssymb}
  - \usepackage{bm}
  - \usepackage{booktabs}
  - \usepackage{fvextra}
  # a Unicode-complete monospace font so the summary()/print() output renders its
  # generators verbatim: the partial-derivative sign and subscripts (X_1 = c_i d/dz_i)
  # are absent from the default Latin Modern Mono.
  - \usepackage{fontspec}
  - \setmonofont{DejaVu Sans Mono}[Scale=0.82]
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,breakanywhere,commandchars=\\\{\}}
vignette: >
  %\VignetteIndexEntry{Structural identifiability in dMod}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  echo     = TRUE
)
# symmetryDetection() is purely symbolic: it drives a Python module (sympy,
# numpy, scipy) through reticulate and needs no compiled model. Every
# computational chunk runs only if sympy is already present, so building the
# vignette never triggers a Python install.
haveSympy <- isTRUE(tryCatch(
  reticulate::py_module_available("sympy"),
  error = function(e) FALSE))
```

\setcounter{tocdepth}{3}
\tableofcontents
\newpage

# Introduction

A parameter is *structurally non-identifiable* when the model produces
exactly the same output for a whole family of parameter values, so that
no amount of data, and no reduction of measurement noise, can tell those
values apart. This is a property of the model equations and the
observation map alone, before any data is collected. It is distinct from
*practical* non-identifiability, which depends on the data at hand and on
noise, and which therefore cannot be decided from the model structure.
`symmetryDetection()` addresses the structural question only.

Concretely, consider an ODE model with states $x \in \mathbb{R}^n$ and
parameters $\theta \in \mathbb{R}^q$,
\begin{equation}
\dot x(t) = f(x(t), \theta), \qquad x(0) = x_0(\theta),
\label{eq:ode}
\end{equation}
together with an observation function
\begin{equation}
g : (x, \theta) \longmapsto g(x, \theta),
\label{eq:obs}
\end{equation}
which collects the quantities that are actually measured, for example a
scaled or offset readout of one or several states. Evaluated along a
solution, $g\big(x(t), \theta\big)$ is the noise-free model output; the
data adds measurement noise,
\begin{equation}
y^{\mathrm{data}}(t) = g\big(x(t), \theta\big) + \eta(t).
\end{equation}
The structural identifiability question concerns $g$ alone, before any
data: do two different unknown vectors produce the same output
$g\big(x(t), \theta\big)$ for all $t$? If they do, the data constrains only
certain combinations of the unknowns, and the remaining directions are
free.

dMod answers this through the `method` argument of `symmetryDetection()`,
whose four settings fall into two families. One tests identifiability by a
rank computation and names the unconstrained combinations
(`"observability"`); the others return the explicit symmetry generators, in
full generality (`"polynomial"`) or as the exact multiplicative and additive
lattices that cover the bulk of practical cases (`"scaling"`,
`"translation"`).

* `method = "observability"` (the default) builds the
  *observability-identifiability matrix* and reads off, from its rank,
  whether the model is locally identifiable and, if not, which
  combinations of states and parameters are unconstrained. It uses no
  ansatz and, for a rational model, is evaluated entirely in exact
  finite-field arithmetic, so it scales to large systems. With
  `reconstruct = TRUE` it returns each non-identifiable direction not only
  by its support but as an exact rational function of the parameters
  (Section \ref{sec-closedform}).
* `method = "polynomial"` searches for *continuous symmetries* of the model
  via the polynomial Lie-symmetry ansatz and returns the explicit
  transformation that leaves the output invariant. This is the object a
  modeller can fold into a parameter transformation to remove the redundancy.
* `method = "scaling"` returns the scaling (toric) symmetries
  $z_i \mapsto \lambda^{c_i} z_i$ exactly, from the integer kernel of the
  monomial-exponent conditions. It is pure integer linear algebra and so
  scales to large models, but it sees only the scaling part of the
  symmetry algebra (Section \ref{sec-scaling}).
* `method = "translation"` returns the additive (constant-shift) symmetries
  $z_i \mapsto z_i + \varepsilon\, a_i$ exactly, the additive analogue of
  `"scaling"`, from the same integer/finite-field linear algebra.

Sections \ref{sec-sym} and \ref{sec-oim} develop the mathematics of the two
methods from first principles, assuming only multivariable calculus.
Section \ref{sec-supply} shows how the methods consume the dMod pipeline objects
`g`, `x`, and `p`. Section \ref{sec-examples} works through examples, and
Section \ref{sec-scope} states the scope and the limitations.

## Structural identifiability, local and global

The terms used throughout are defined as follows. Collect the unknowns
into a single vector $z = (x_0, \theta)$, the initial state and the
parameters, and write $g(t\,; z) := g\big(x(t; z), \theta\big)$ for the
noise-free output that $z$ produces through \eqref{eq:ode} and the
observation function \eqref{eq:obs}. Two unknown vectors are
*output-equivalent* when they produce the same output for all admissible
times (and, where the model has inputs, all admissible inputs),
\begin{equation}
z \sim z' \quad :\Longleftrightarrow \quad
g(t\,; z) = g(t\,; z') \ \text{ for all } t.
\end{equation}
Output-equivalent vectors are indistinguishable from data, with or without
noise.

A component $z_i$ is *structurally globally identifiable* if, for a
generic $z$, output-equivalence forces that component to agree,
\begin{equation}
z \sim z' \;\Longrightarrow\; z_i = z_i'.
\label{eq:sgi}
\end{equation}
It is *structurally locally identifiable* if \eqref{eq:sgi} holds only
after restricting $z'$ to a neighbourhood of $z$: there is a neighbourhood
$U \ni z$ such that
\begin{equation}
z' \in U \ \text{and}\ z \sim z' \;\Longrightarrow\; z_i = z_i'.
\label{eq:sli}
\end{equation}
The model is structurally globally (respectively locally) identifiable
when every component is. A component is *structurally locally
non-identifiable* when \eqref{eq:sli} fails: every neighbourhood of $z$
contains a distinct output-equivalent $z'$, so even perfect, noise-free
data cannot pin the component down. It is *structurally globally
non-identifiable* when \eqref{eq:sgi} fails: some $z' \neq z$, not
necessarily close to $z$, reproduces the output. Local non-identifiability
implies global non-identifiability; global is the stronger requirement,
because it additionally excludes *isolated* alternative values far from
$z$.

The two failure modes map onto the two methods.

* A *continuous* family of output-equivalent vectors through $z$ is a
  local non-identifiability. Its tangent at $z$ is a null vector of the
  observability-identifiability matrix of Section \ref{sec-oim}, and its
  explicit form along the whole family is the symmetry generator of
  Section \ref{sec-sym}. Such a
  non-identifiability is local and, traced along its orbit, also global.
* An *isolated* output-equivalent vector $z' \neq z$ (a discrete
  symmetry, for example a sign flip) is purely global and has no
  infinitesimal signature. Neither method targets it: a full-rank
  observability matrix certifies *local* identifiability only, and the
  polynomial search returns *continuous* generators. Ruling out discrete
  symmetries requires a separate, global argument.

# Symmetries and the determining equations {#sec-sym}

## A transformation that the data cannot see

Group the unknowns into a single vector
$z = (x_1, \dots, x_n, \theta_1, \dots, \theta_q) \in \mathbb{R}^{n+q}$.
Parameters are constants of the motion, so the augmented system carries
$\dot\theta = 0$ alongside \eqref{eq:ode}. A *symmetry* is a smooth,
invertible map $z \mapsto \Phi_\varepsilon(z)$, depending on a real
parameter $\varepsilon$ with $\Phi_0 = \mathrm{id}$, that

1. maps every solution of the dynamics to another solution, and
2. leaves the output \eqref{eq:obs} unchanged.

If such a family exists for a continuous range of $\varepsilon$, then the
unknowns can be moved along it without changing the output
$g\big(x(t), \theta\big)$. That is precisely
a structural non-identifiability, and $\Phi_\varepsilon$ is its explicit
form. The canonical example is a scaling that trades a state against a
calibration parameter: if only $\alpha\, x_1$ is observed, then doubling
$\alpha$ and halving $x_1$ (and whatever else the dynamics couples to
$x_1$) reproduces the same readout.

## The infinitesimal generator

Working with the finite map $\Phi_\varepsilon$ directly is awkward.
Instead one differentiates it at $\varepsilon = 0$ and works with the
resulting velocity field
\begin{equation}
\xi(z) \;=\; \left.\frac{\partial}{\partial \varepsilon}
            \Phi_\varepsilon(z)\right|_{\varepsilon = 0}
        \;=\; \big(\xi_1(z), \dots, \xi_{n+q}(z)\big).
\end{equation}
The vector $\xi(z)$ points in the direction the family moves the point
$z$. The finite transformation is recovered by following this direction,
that is by solving $\frac{d}{d\varepsilon}\Phi_\varepsilon = \xi(\Phi_\varepsilon)$.
It is convenient to package $\xi$ as the first-order differential operator
\begin{equation}
X \;=\; \sum_{i=1}^{n+q} \xi_i(z)\, \frac{\partial}{\partial z_i},
\end{equation}
called the *infinitesimal generator*. For any smooth function $h(z)$, the
quantity
\begin{equation}
X(h) \;=\; \sum_{i=1}^{n+q} \xi_i(z)\, \frac{\partial h}{\partial z_i}
\label{eq:lie-h}
\end{equation}
is the rate of change of $h$ as one moves along the family. It is just the
directional derivative of $h$ in the direction $\xi$. Readers who have met
the *Lie derivative* will recognise \eqref{eq:lie-h} as exactly that;
readers who have not lose nothing by reading $X(h)$ as a directional
derivative.

## Recovering the transformation by integrating the generator {#sec-flow}

Both methods return the generator $\xi$, not the finite map directly: the
observability null vector and the `"polynomial"` solution are the values, or the
field, of $\xi$. The transformation $\Phi_\varepsilon$ that $\xi$ stands for is
recovered from it as the flow of the autonomous ODE
\begin{equation}
\frac{d}{d\varepsilon}\,\Phi_\varepsilon(z) = \xi\big(\Phi_\varepsilon(z)\big),
\qquad \Phi_0(z) = z,
\label{eq:flow}
\end{equation}
the time-$\varepsilon$ flow $\Phi_\varepsilon = \exp(\varepsilon X)$ of the vector
field $X$. Equivalently each coordinate evolves by
$dz_i/d\varepsilon = \xi_i\big(z(\varepsilon)\big)$. The reported direction is the
velocity at $\varepsilon = 0$; the finite transformation is one integration away.

For a *scaling* the generator has constant weights $\xi_i = c_i z_i$, so
\eqref{eq:flow} decouples into $dz_i/d\varepsilon = c_i z_i$ and integrates at
once to $z_i(\varepsilon) = z_i(0)\,e^{c_i \varepsilon}$. With
$\lambda = e^\varepsilon$ this is the monomial rescaling
\begin{equation}
z_i \;\longmapsto\; \lambda^{c_i}\, z_i ,
\label{eq:scaling-flow}
\end{equation}
which is why a scaling is reported by its integer weight vector $c$: those weights
*are* the generator, and the finite symmetry is the power law
\eqref{eq:scaling-flow} they integrate to. Halving a state while doubling its
calibration parameter, the example above, is \eqref{eq:scaling-flow} at
$\varepsilon = \pm\log 2$.

For a *general* direction the components $\xi_i$ are rational functions and
\eqref{eq:flow} is a genuinely nonlinear, coupled flow. It still defines a
one-parameter family $\Phi_\varepsilon$ through $z$, a curved orbit rather than a
ray, but its integral need not be elementary. What the method delivers is always
$\xi$, exactly; $\Phi_\varepsilon$ follows by integrating it, in closed form when
the flow is simple (the hyperbola $k_{tx}k_{tl} = \text{const}$ of the
gene-expression example below) and otherwise as a well-defined flow described
through the quantities it leaves fixed. In every case the orbit stays on the level
sets of the observed combinations, which is exactly the content of $X(g) = 0$.

## The two determining equations

The two defining properties of a symmetry translate into two conditions
on $\xi$.

*Output invariance.* The output must not change along the family. With
$h = g$ in \eqref{eq:lie-h},
\begin{equation}
X(g) \;=\; \sum_{i=1}^{n+q} \xi_i\, \frac{\partial g}{\partial z_i} \;=\; 0.
\label{eq:obs-cond}
\end{equation}

*Compatibility with the dynamics.* The family must send solutions to
solutions. Write the right-hand side of \eqref{eq:ode} as the vector field
$F = \sum_{i=1}^{n} f_i\, \partial / \partial x_i$. Mapping a point first
and then flowing it along $F$ must give the same result as flowing first
and then mapping. The infinitesimal version of this statement is that the
two operators commute, $[F, X] = 0$, where the *commutator* (Lie bracket)
has the components
\begin{equation}
[F, X]_i \;=\; \sum_{l=1}^{n}
  \left( f_l\, \frac{\partial \xi_i}{\partial x_l}
        \;-\; \xi_l\, \frac{\partial f_i}{\partial x_l} \right) \;=\; 0,
\qquad i = 1, \dots, n.
\label{eq:flow-cond}
\end{equation}
The first term is the change of $\xi_i$ as one flows along $F$; the second
is the change of $f_i$ as one moves along $\xi$. Their cancellation is the
statement that the transformation and the flow do not interfere.

Equations \eqref{eq:obs-cond} and \eqref{eq:flow-cond} together are the
*determining equations*. Every $\xi$ that satisfies both generates a
structural non-identifiability.

## Solving the determining equations

The determining equations are linear in $\xi$, but $\xi$ is an unknown
function. dMod restricts the search to polynomial $\xi$: each component
$\xi_i$ is an unknown polynomial in the variables $z$ up to a chosen
degree `pMax`,
\begin{equation}
\xi_i(z) \;=\; \sum_{|\alpha| \le \texttt{pMax}} c_{i,\alpha}\, z^{\alpha},
\end{equation}
with unknown coefficients $c_{i,\alpha}$. Substituting this ansatz into
\eqref{eq:obs-cond} and \eqref{eq:flow-cond} and clearing denominators
gives polynomial identities in $z$. A polynomial is zero only if every one
of its monomial coefficients is zero, so each identity becomes a set of
linear equations in the unknowns $c_{i,\alpha}$. The symmetries are the
null space of this linear system. dMod assembles and solves it in exact
rational arithmetic (a modular Gaussian elimination with rational
reconstruction, validated by back-substitution), so the rank decision that
controls how many symmetries are reported is exact rather than a
floating-point threshold. Each surviving generator is then verified
symbolically by checking \eqref{eq:obs-cond} and \eqref{eq:flow-cond}
directly.

The degree `pMax`, the ansatz type, the extra Lie-derivative order, the symbolic
backend and the verification flag are the fields of `polynomialControl()`, passed as
`symmetryDetection(..., method = "polynomial", polynomial = polynomialControl(...))`.
For example `polynomialControl(ansatz = "uni", pMax = 1L)` restricts the search to
univariate linear infinitesimals.

## Observables with rational gradient

Only the *gradient* of $g$ enters the output condition
\eqref{eq:obs-cond}; the value of $g$ never does. Consequently $g$ may be
non-rational as long as each $\partial g / \partial z_i$ is rational. The
practically important case is a measurement on logarithmic scale: for
$g = \log_{10}(x_1) + c$,
\begin{equation}
\frac{\partial g}{\partial x_1} = \frac{1}{x_1 \ln 10},
\qquad
\frac{\partial g}{\partial c} = 1,
\end{equation}
both rational, so the determining equations are well posed even though
$g$ itself is transcendental. dMod differentiates $g$ first and only then
builds the polynomial system, so observables such as `log10(x1) + c` are
handled by `"polynomial"` without special treatment. An observation whose
gradient is not rational, for example $\exp(x_1)$ or $\sqrt{x_1}$, is
outside the reach of both methods. The `"observability"` method of the next
section is stricter still: its finite-field construction needs the value of
$g$, not only its gradient, to be rational, so a logarithmic readout is
given to it in the equivalent rational form discussed there.

# The observability-identifiability matrix {#sec-oim}

The default method, selected by `method = "observability"`, takes a
different route to the same information and needs no ansatz. It is the
local algebraic observability test of Sedoglavic (2002), extended so that
the unknowns include the parameters and the initial states, which is the
full input-state-parameter observability (FISPO) setting of Villaverde et
al. (2016) and the basis of their STRIKE-GOLDD toolbox. The construction
below is referred to throughout as the observability-identifiability
matrix (OIM).

## Output derivatives are functions of the unknowns

At $t = 0$ the output is the observation function $g$ itself. Its time
derivative follows by the chain rule and the dynamics \eqref{eq:ode},
\begin{equation}
\frac{d}{dt}\, g\big(x(t), \theta\big)\Big|_{t=0}
   = \sum_{i=1}^{n} \frac{\partial g}{\partial x_i}\, f_i
   \;=\; L_F\, g,
\end{equation}
where $L_F\, g$ is the derivative of $g$ along the flow, that is the same
directional derivative as before but now taken in the direction $F$.
Iterating, the $k$-th time derivative of the output at $t = 0$ is
$L_F^{\,k} g$. The sequence
\begin{equation}
\big(\, g,\; L_F g,\; L_F^2 g,\; \dots,\; L_F^N g \,\big)
\label{eq:summary}
\end{equation}
is the list of Taylor coefficients of the output $g\big(x(t), \theta\big)$
at $t = 0$. It is everything the output curve can reveal about the unknowns
near $t = 0$, and each entry is an explicit function of $z$. Matching output
Taylor coefficients to identify parameters is the classical power-series method of
Pohjanpalo (1978), used numerically in tools such as SIAN (Hong et al., 2020).

## Identifiability as injectivity

Collect \eqref{eq:summary} into a single map
$\Phi: z \mapsto (g, L_F g, \dots, L_F^N g)(z)$. If two unknown vectors
give the same output curve, they give the same values for all the
derivatives, hence the same $\Phi(z)$. Locally, the unknowns are therefore
determined exactly when $\Phi$ is injective near a generic point, which
holds when its Jacobian
\begin{equation}
\mathcal{O}(z) \;=\; \frac{\partial \Phi}{\partial z}
\;=\;
\begin{pmatrix}
\partial g / \partial z \\[2pt]
\partial (L_F g) / \partial z \\[2pt]
\vdots \\[2pt]
\partial (L_F^N g) / \partial z
\end{pmatrix}
\label{eq:oim}
\end{equation}
has full column rank $n + q$. The matrix $\mathcal{O}$ in \eqref{eq:oim} is
the observability-identifiability matrix (OIM). The rank verdict,
\begin{equation}
\operatorname{rank} \mathcal{O} = n + q
\;\Longleftrightarrow\;
\text{structurally locally identifiable,}
\end{equation}
is the rank condition of Sedoglavic (2002); including the parameters and
the initial states among the columns of $\mathcal{O}$ is the FISPO
extension of Villaverde et al. (2016).
When the rank is deficient, the *null space* of $\mathcal{O}$ collects the
directions in $z$ along which the output and all its derivatives are
unchanged. These directions are exactly the values $\xi(z)$ of the
infinitesimal generators of Section \ref{sec-sym} at the point $z$: the two methods
describe the same object, one as a tangent vector at a point, the other as
an explicit field. Recovering Lie-symmetry directions from this observability
setting is the subject of Massonis and Villaverde (2020).

## The matrix on a small model

A single example makes the construction concrete. Take one species with a
lumped influx and linear decay, measured directly,
\begin{equation}
\dot A \;=\; k_1 + u\, k_2 - k_d\, A, \qquad y \;=\; A,
\label{eq:toy-oim}
\end{equation}
where $u$ is a known, constant input. The unknowns are
$z = (A_0, k_1, k_2, k_d)$, the initial value and the three rates. The output
and its derivatives along the flow are read straight off \eqref{eq:toy-oim}:
\begin{align}
g          &= A, \nonumber\\
L_F g      &= k_1 + u\, k_2 - k_d\, A, \nonumber\\
L_F^2 g    &= -k_d\,(k_1 + u\, k_2) + k_d^2\, A,
\label{eq:toy-lie}
\end{align}
each obtained by differentiating the previous line and substituting $\dot A$
once more. Differentiating \eqref{eq:toy-lie} in $z = (A_0, k_1, k_2, k_d)$, and
evaluating at $t = 0$ where $A = A_0$, gives the first three rows of the OIM,
\begin{equation}
\mathcal{O} \;=\;
\begin{pmatrix}
1            & 0     & 0     & 0 \\
-k_d         & 1     & u     & -A_0 \\
k_d^2        & -k_d  & -u\,k_d & -(k_1 + u\,k_2) + 2 k_d A_0
\end{pmatrix}.
\label{eq:toy-matrix}
\end{equation}
The two rates enter every derivative only through the combination
$K = k_1 + u\, k_2$: the influx cannot be split into its baseline $k_1$ and its
input-driven part $k_2$ from the output alone. The matrix \eqref{eq:toy-matrix}
therefore has a one-dimensional null space, spanned by the vector that trades
$k_1$ against $k_2$ at fixed $K$,
\begin{equation}
\xi \;=\; \frac{\partial}{\partial k_2} - u\,\frac{\partial}{\partial k_1},
\qquad\text{that is}\qquad
\xi_{k_1} = -u, \quad \xi_{k_2} = 1, \quad \xi_{A_0} = \xi_{k_d} = 0 .
\label{eq:toy-null}
\end{equation}
One checks $\mathcal{O}\,\xi = 0$ row by row: the $g$ row is $0$, the $L_F g$ row
is $-u + u = 0$, and the $L_F^2 g$ row is $u\, k_d - u\, k_d = 0$; the higher
derivatives add no new constraint because they too depend on the rates only
through $K$, which is exactly why the Lie order saturates. The remaining
coordinates $A_0$, $K$ and $k_d$ are identifiable. This direction
\eqref{eq:toy-null} is what the method returns, as the exact constant direction
`k1 : u, k2 : -1` (the same line \eqref{eq:toy-null} up to its overall sign,
which the gauge fixes); carrying $\mathcal{O}$ symbolically is
feasible here but not on a network of dozens of states, which is the subject of
the next sections. Note also that $\xi$ depends on the input value $u$: holding
$u$ at one generic value reports this trade-off, whereas measuring at two
distinct values of $u$ pins $k_1$ and $k_2$ separately (Section \ref{sec-conditions}).

## Evaluation over a finite field {#sec-finitefield}

Carrying $\mathcal{O}$ as a symbolic matrix is what makes the naive test
intractable: the entries $L_F^{\,k} g$ are rational functions whose
numerators and denominators grow with every Lie derivative, and the
expression swell, not the linear algebra, dominates the cost. The remedy,
following Sedoglavic (2002), is to evaluate $\mathcal{O}$ at a single
generic point and to reduce the resulting numerical matrix, because the rank
at a generic point equals the rank of the symbolic matrix away from a set of
measure zero.

dMod performs that evaluation in exact arithmetic over a finite field
$\mathrm{GF}(p)$, the integers modulo a prime $p$. There are two reasons to
work modulo a prime rather than over the rationals. First, exactness without
swell: every intermediate number stays in $\{0, 1, \dots, p-1\}$, so the
Gaussian elimination that computes the rank never produces the growing
fractions that exact rational arithmetic would, yet the result is exact, not
rounded. Second, $p$ is prime precisely so that $\mathrm{GF}(p)$ is a
*field*: every nonzero residue has a multiplicative inverse, which is what
row reduction needs to create pivots. A composite modulus would have
zero divisors and the elimination could break down. dMod uses primes just
below $2^{31}$, so that products of two residues stay within a $63$-bit
integer and the modular arithmetic is exact machine arithmetic.

For a rational model $\mathcal{O}$ is never written symbolically at all. Its
rows are the gradients of the Taylor coefficients of the output, and both the
coefficients and their gradients are produced numerically by integrating the
state as a truncated power series in time.

Expand the solution at the evaluation point as a power series in $t$,
\begin{equation}
x(t) = \sum_{k \ge 0} x_k\, t^k, \qquad x_0 = x(0),
\end{equation}
and the right-hand side along that solution as
$f\big(x(t)\big) = \sum_{k \ge 0} \phi_k\, t^k$. Matching the two sides of
$\dot x = f(x)$ coefficient by coefficient gives the recurrence
\begin{equation}
(k+1)\, x_{k+1} = \phi_k,
\qquad\text{that is}\qquad
x_{k+1} = \tfrac{1}{k+1}\, \phi_k .
\label{eq:taylor-rec}
\end{equation}
The coefficient $\phi_k$ depends only on $x_0, \dots, x_k$, because the power
series of a sum, a product or a reciprocal is built from lower orders: for a
product $c = a\,b$ the coefficient is a convolution
$c_k = \sum_{i=0}^{k} a_i\, b_{k-i}$, and for a reciprocal $c = 1/a$ the same
convolution rearranged,
$c_k = a_0^{-1}\big(\delta_{k0} - \sum_{i=1}^{k} a_i\, c_{k-i}\big)$. Running the
sequence of operations that defines $f$ on these truncated series therefore
yields $\phi_0, \phi_1, \dots$, and \eqref{eq:taylor-rec} fills
$x_1, x_2, \dots$ one order at a time. This is automatic differentiation in
*Taylor mode*: no symbolic expression for $L_F^{\,k} g$ is ever formed, yet the
coefficient of $t^k$ in $g\big(x(t)\big)$ equals $L_F^{\,k} g / k!$, the $k$-th
output derivative.

The columns of $\mathcal{O}$ are the derivatives with respect to the unknowns
$z$. They are obtained in the same pass by carrying, next to every series
coefficient, its gradient with respect to $z$: each number $a$ in the
computation is replaced by the pair $\big(a,\, \partial a / \partial z\big)$,
and the arithmetic is extended by the product rule
$\partial(ab) = a\,\partial b + b\,\partial a$ and its analogues. This is
forward-mode dual-number differentiation, nested inside the Taylor recurrence;
after the recurrence the gradient parts of the coefficients of $g\big(x(t)\big)$
are exactly the rows of $\mathcal{O}$.

Everything runs in $\mathrm{GF}(p)$, and each coefficient at each order is
computed once, so the work is a few convolutions per operation and grows only
polynomially with the model size and the number of derivatives. Deep chains
that defeat a symbolic construction are handled in a fraction of a second. The
number of derivatives $N$ is raised until the rank stops growing, which is
guaranteed by $N = n + q - 1$, so the order need not be set by hand.

Because the construction lives over a field built from $+, -, \times, \div$
only, it is defined exactly for *rational* right-hand sides and observables;
a non-rational input raises an error, the one exception being a power or Hill
term $\mathrm{base}^{\,n}$ with a parameter exponent, which is recast to a
rational coordinate (with or without `equilibrate`; Section \ref{sec-supply}). A logarithmic readout
$\log_{10}(h) + \mathrm{offset}$, common in dMod models, carries the same
information as the rational observable $\mathrm{scale}\cdot h$: a monotone
transformation has the same level sets, and the additive offset is itself
identifiable and plays the role of a multiplicative scale. It is supplied to
`"observability"` in that rational form (or analysed with `"polynomial"`, which
differentiates first and so accepts the rational gradient of $\log_{10}$).

A single prime and a single point already give the rank with probability
one, but the verdict does not rest on one draw. dMod repeats the evaluation
at several deterministic generic points and over several primes and takes the
maximal rank; agreement certifies it, since a non-generic coincidence at
every point and prime has probability zero. The support of each null vector,
the set of states and parameters with a nonzero entry, names the unknowns
that take part in the non-identifiability. With `reconstruct = TRUE` the
directions are returned not merely by support but as exact rational functions
of the parameters, by the reconstruction of Section \ref{sec-closedform}.

The remaining price is that the verdict is *local*: it concerns
identifiability around generic points, and a null vector gives the tangent
direction rather than the finite transformation. When the explicit finite
transformation is wanted, the `"polynomial"` method returns it.

## Scaling symmetries and large models {#sec-scaling}

Two devices give exact directions on models too large for the general
closed form, by exploiting that real non-identifiable directions are
structured.

The dominant class is scaling: $z_i \mapsto \lambda^{c_i} z_i$, whose
generator $X = \sum_i c_i z_i\, \partial / \partial z_i$ has a constant
tangent, the integer vector $c$. The two conditions of Section \ref{sec-sym} reduce to
integer-linear ones. Output invariance forces every monomial exponent of
the observation to be orthogonal to $c$; compatibility with the dynamics
forces every monomial of $f_i$ to carry the weight of $x_i$, that is
$(b - e_i) \cdot c = 0$ for each exponent $b$ of $f_i$. Stacking these as an
integer matrix and taking its kernel returns all scaling symmetries
exactly, with no point evaluation and no expression swell, so
`method = "scaling"` scales to large models. It sees only the scaling part
of the symmetry algebra, which in practice is the bulk of identifiability
problems (units, calibration, absolute levels). The symbolic backend and the
evaluation point are the fields of `scalingControl()`, passed as
`symmetryDetection(..., method = "scaling", scaling = scalingControl(...))`.

## Closed-form directions by multi-point reconstruction {#sec-closedform}

The rank verdict needs $\mathcal{O}$ at a single point. The *functional*
form of a non-identifiable direction needs more, and `reconstruct = TRUE`
recovers it by sampling. Over the field of rational functions
$\mathbb{Q}(z)$ the null space of $\mathcal{O}$ has a basis whose entries are
themselves rational functions of $z$; evaluating $\mathcal{O}$ at one point
yields one numerical null vector, that is a single *sample* of each of those
functions. Reconstructing a function from samples is interpolation, and the
whole subtlety is to do it exactly.

**A canonical sample.** At a point $z^\ast$ the reduced row echelon form of
$\mathcal{O}(z^\ast)$ over $\mathrm{GF}(p)$ splits the coordinates into pivot
and free columns. Each free column $f$ gives one basis vector $v$ of the null
space, normalised so that $v_f = 1$ and $v_c = -R_{c f}$ on the pivot
coordinates $c$, where $R$ is the reduced matrix. This normalisation is
canonical: as long as the pivot pattern is the generic one, the same
free column $f$ yields the *same* rational functions $r_c(z) := v_c$ at every
point, so the values obtained at different points are samples of one fixed
function and may be interpolated entry by entry.

**Two reductions.** First, the scaling directions are removed by the integer
kernel of Section \ref{sec-scaling}; they already have an exact closed form, so
only the residual, non-scaling directions are reconstructed. Across several
conditions the scalings are those common to every condition, the kernel of the
stacked monomial-exponent conditions, so a scaling is obtained exactly whether the
analysis runs on one condition or many. Such a scaling is reported by its
integer toric weights $c_i$ (the generator $z_i \mapsto \lambda^{c_i} z_i$),
exactly and regardless of `reconstruct`; only the non-scaling residue is sampled.
Second, each *entry* of a residual direction is probed for its *relevant*
variables by perturbing each coordinate of the evaluation point in turn and
keeping those that move that entry; the entry is then reconstructed over that
small set $u$ only, the remaining coordinates held at generic values. Because the
probe is per entry, a direction may couple many coordinates while each of its
entries stays narrow, and the cost is set by the widest single entry, not by the
direction or the model. When a known input is left as a single generic symbol,
the probed coordinates include it, and a direction may come back as a rational
function of that input; this is usually a sign that the analysis should be split
across the conditions in which the input takes its actual values
(Section \ref{sec-conditions}), after which the input is a constant of each
problem and no longer appears. An entry that depends on no variable is a constant
rational and is read off directly.

**The per-entry fit.** Write each pivot entry as
$r_c(u) = \mathrm{num}(u)/\mathrm{den}(u)$ with numerator and denominator
polynomials of total degree at most $D$ in the relevant variables. Clearing
the denominator gives, at every sample point $u^{(s)}$,
\begin{equation}
\mathrm{num}\!\left(u^{(s)}\right)
\;-\; r_c\!\left(u^{(s)}\right)\,\mathrm{den}\!\left(u^{(s)}\right) \;=\; 0 ,
\label{eq:fit}
\end{equation}
which is *linear and homogeneous* in the unknown coefficients of $\mathrm{num}$
and $\mathrm{den}$. Stacking \eqref{eq:fit} over the sample points gives a
matrix whose kernel is the coefficient vector, determined up to one common
scale that cancels in the ratio $\mathrm{num}/\mathrm{den}$. Solving the
homogeneous system, rather than fixing the constant term of $\mathrm{den}$ to
one, is what lets the fit recover a denominator with no constant term, a bare
rate constant for instance. The degree $D$ is raised from zero: below the
true degree the only solution is trivial (an inconsistent fit), at the true
degree the kernel is exactly one dimensional, and that minimal $D$ is taken.
Sampling a few points beyond the number of unknowns makes a spurious
low-degree fit fail \eqref{eq:fit} on the extra points, so it cannot be
accepted by accident.

**Wide entries: sparse interpolation.** The dense fit costs
$\binom{n_{\mathrm{rel}}+D}{n_{\mathrm{rel}}}$ samples, which is prohibitive when
a single entry couples many variables. The wide entries that occur in practice
are the conserved-moiety cofactors, a sum over many species divided by a single
rate, $-(x_1+\dots+x_k)/r$. Such an entry is a *Laurent polynomial*, and on the
geometric schedule $u^{(k)} = (\beta_1^{k},\dots,\beta_n^{k})$ its values form a
linear-recurrence sequence, so it is recovered by Ben-Or-Tiwari interpolation
(Ben-Or and Tiwari, 1988): the Berlekamp-Massey algorithm gives the number of
terms and a locator
polynomial whose roots are the monomial values $\prod_j \beta_j^{e_j}$, the
exponents are read back from those values, and a transposed-Vandermonde solve
gives the coefficients. The work is set by the number of terms, not by
$\binom{n_{\mathrm{rel}}+D}{n_{\mathrm{rel}}}$, so a sparse entry in many
variables is recovered from a handful of samples.

An entry with a genuine multi-term denominator is not itself a linear-recurrence
sequence, so Ben-Or-Tiwari cannot be applied to it directly. It is handled by
separating numerator and denominator first. From a generic shift $s$, the entry
along the ray $s + t\,(u-s)$ is a univariate rational in $t$; a Cauchy fit
normalised so the denominator has constant term one evaluates $N(u)/D(s)$ and
$D(u)/D(s)$ at any $u$, where the gauge $D(s)$ is a constant that cancels in
$N/D$. Both are now ordinary sparse polynomials, each recovered by Ben-Or-Tiwari
on the geometric schedule and lifted across primes. A direction whose numerator
or denominator exceeds the degree caps, or is too dense to be sparse, is not
reconstructed and falls back to its support.

**Certifying the result.** Reconstruction is exact arithmetic, but to guard
against an unlucky evaluation or a non-generic pivot pattern, every reconstructed
direction is re-evaluated at a fifth prime, disjoint from the four used to build
it, and checked against the null vector there. A direction that fails this check
is reported by its support with `reconstruct = FALSE` rather than as a wrong
closed form, so the closed forms that are returned are certified.

**Why primes: lifting $\mathrm{GF}(p)$ back to $\mathbb{Q}$.** Each sample of
$r_c$ and each kernel in \eqref{eq:fit} is computed in $\mathrm{GF}(p)$, for
the reasons of Section \ref{sec-finitefield}: exactness without expression swell. A single
prime therefore returns the coefficients of $\mathrm{num}$ and $\mathrm{den}$
only as residues modulo $p$, not as the rational numbers we want. They are
recovered in two exact steps. Compute the coefficients modulo several primes
$p_1, \dots, p_r$. The *Chinese Remainder Theorem* combines the residues of
one coefficient into its value modulo the product $M = p_1 \cdots p_r$: there
is a unique $x \in \{0, \dots, M-1\}$ with $x \equiv x_i \pmod{p_i}$ for every
$i$. *Rational reconstruction* then turns that integer back into a fraction:
the extended Euclidean algorithm applied to $x$ and $M$ returns the unique
$n/d$ with
\begin{equation}
x \,\equiv\, n\, d^{-1} \pmod{M},
\qquad |n|,\ |d| \,\le\, \sqrt{M/2},
\label{eq:ratrec}
\end{equation}
whenever such a bounded fraction exists. The bound in \eqref{eq:ratrec} is
the point of using several primes: the true coefficients of an identifiability
direction are small integers or simple fractions, so once $M$ is large enough
that $\sqrt{M/2}$ exceeds them, the reconstruction is exact. dMod uses four
primes just below $2^{31}$, giving $M \approx 2^{124}$ and a reconstruction
range up to $|n|, |d| \approx 2^{61}$, far beyond anything these coefficients
reach. Disagreement between primes, finally, certifies the lift: an accidental
or non-generic value at one prime does not reproduce across the others.

**A worked reconstruction.** Suppose one coefficient of a direction is the
rational number $-\tfrac{1}{3}$, and the fit has returned it only as a residue
modulo two (small, for illustration) primes $p_1 = 7$ and $p_2 = 11$. Modulo a
prime, division is multiplication by the modular inverse: $3^{-1} \equiv 5
\pmod 7$ since $3\cdot 5 = 15 \equiv 1$, so $-\tfrac{1}{3} \equiv -5 \equiv 2
\pmod 7$; likewise $3^{-1} \equiv 4 \pmod{11}$ gives $-\tfrac{1}{3} \equiv -4
\equiv 7 \pmod{11}$. The fit thus hands us the two residues $2 \pmod 7$ and
$7 \pmod{11}$, with no sign of the fraction they came from. The Chinese
Remainder Theorem combines them into one residue modulo $M = 7\cdot 11 = 77$:
the unique $x \in \{0,\dots,76\}$ with $x \equiv 2 \pmod 7$ and $x \equiv 7
\pmod{11}$ is $x = 51$. Rational reconstruction then runs the extended Euclidean
algorithm on $(77, 51)$, stopping once the remainder drops to or below the bound
$\lfloor\sqrt{M/2}\rfloor = 6$,
\begin{equation}
\begin{array}{r|rrrrr}
r & 77 & 51 & 26 & 25 & 1 \\
s & 0  & 1  & -1 & 2  & -3
\end{array}
\label{eq:ratrec-worked}
\end{equation}
where each column subtracts one quotient-multiple of the next, in $r$ and in the
cofactor $s$ alike. The first remainder at or below $6$ is $r = 1$ with cofactor
$s = -3$, giving $n/d = 1/(-3) = -\tfrac{1}{3}$ after fixing the sign of the
denominator. Both $|n| = 1$ and $|d| = 3$ are within the bound, so the lift is
accepted, and the residues $2 \pmod 7$, $7 \pmod{11}$ have been turned back into
$-\tfrac{1}{3}$ exactly. With the four primes near $2^{31}$ the same procedure
recovers any coefficient with numerator and denominator up to about $2^{61}$.

**Guarantees and cost.** Every step, the samples, the kernels, the Chinese
remaindering and the reconstruction, is exact integer or finite-field
arithmetic; no floating-point number enters and no rounding threshold is
chosen. A direction whose relevant set or whose minimal degree exceeds a cap
is returned with its support and `reconstruct = FALSE` rather than a guess, so
the call never stalls on expression swell. Those caps, together with the
saturation and gap-order bounds, are the fields of `reconstControl()`, passed as
`symmetryDetection(..., control = reconstControl(...))`; raising them recovers
wider or higher-degree directions at the cost of more samples. The per-point
construction of $\mathcal{O}$, run once for each sample and prime, is carried out
by the same compiled finite-field kernel as the verdict, which is what makes the
reconstruction feasible on large models. The sample bank for a direction is built
in one batched kernel call that solves every (point, prime) pair at once and
spreads them over `cores` threads, so the reconstruction occupies all cores and
not only the per-condition build does. The closed-form *finite*
transformation within a polynomial ansatz remains the province of the
`"polynomial"` method; the reconstruction here returns the infinitesimal
generator $\xi(z)$ of Section \ref{sec-sym}, exactly, at scale.

## Several experimental conditions {#sec-conditions}

A model is rarely measured under a single setting. The same network is run with
the proteasome inhibited or not, with translation blocked or not, at several
ligand doses, sometimes with a receptor knocked down, and the parameters are
shared across these *conditions* except for the few that the perturbation
changes. Identifiability is then a property of the whole experiment: a parameter
combination is non-identifiable only if no condition can see it.

Each condition $c$ fixes the externally controlled quantities (the boolean
switches, the dose) to known values and may reparametrise a few rates (a
knockdown-specific synthesis rate, say). It therefore has its own right-hand
side $f^{(c)}$ and observation map $g^{(c)}$ over a shared set of unknowns $z$,
the union of the parameters and free initial values appearing in any condition,
and its own observability matrix $\mathcal{O}^{(c)}(z)$. A direction $v$ is
invisible in the whole experiment exactly when it is invisible in every
condition, $\mathcal{O}^{(c)}(z)\,v = 0$ for all $c$, which is the null space of
the stacked matrix
\begin{equation}
\mathcal{O}(z) \;=\;
\begin{bmatrix} \mathcal{O}^{(1)}(z) \\ \vdots \\ \mathcal{O}^{(K)}(z) \end{bmatrix},
\qquad
\ker \mathcal{O} \;=\; \bigcap_{c=1}^{K} \ker \mathcal{O}^{(c)} .
\label{eq:stacked}
\end{equation}
The intersection in \eqref{eq:stacked} can be strictly smaller than any single
$\ker\mathcal{O}^{(c)}$: a combination hidden under one switch setting is often
exposed by another, so stacking conditions resolves non-identifiabilities that
no single condition resolves.

This is also the correct treatment of a known switch. Holding a binary switch at
one generic symbol and reading off the null space finds the directions valid for
*that* generic value; a switch that gates a degradation term, for example, then
produces a direction proportional to it, which holds only in one regime. Setting
the switch to its actual value in each condition and stacking removes such
regime-specific artefacts. Because the switch is a constant of each
$\mathcal{O}^{(c)}$ and not a symbol of $z$, it cannot appear in any
reconstructed direction. The reconstruction of Section \ref{sec-closedform} is unchanged: it runs
on the stacked $\mathcal{O}$, so the closed-form directions are the generators of
the intersection. In dMod the conditions are supplied as an [eventlist] together
with a condition grid, from which `symmetryDetection()` builds one observability
problem per condition.

## Several stimuli in time: piecewise segments {#sec-segments}

The same stacking handles several stimuli applied at different times. An event
after the start time $t_0$ breaks the analytic continuation of the output, so the
post-stimulus timeline is split at the distinct event times into segments
$S_0, S_1, \dots$, and one local jet is built per segment over the shared unknowns
$z$ and stacked exactly as in \eqref{eq:stacked}, a condition spread over time.

The subtlety is the seed of a later segment. To build $S_j$'s jet over
$\mathrm{GF}(p)$ one needs the state $x(\tau_j^+)$ *and* its $z$-Jacobian at the
boundary, i.e. the previous segment's state flowed across the gap $\Delta t_{j-1}$
and mapped by the event. That finite-time nonlinear flow is transcendental in the
gap length and is not a finite-field number, so it cannot be evaluated by plugging
in the numeric gap.

The exact device is to keep the gap length a **formal power-series variable**. The
flow of $S_{j-1}$ over its gap is the state's Taylor series in the elapsed time,
re-read as a series in $\Delta t_{j-1}$,
\begin{equation}
x(\tau_j^-) \;=\; \sum_{m\ge 0} x_m\,\Delta t_{j-1}^{\,m},
\label{eq:gapseries}
\end{equation}
whose coefficients $x_m$ are exactly the Taylor coefficients the kernel already
computes (with their $z$-duals). The event map composes on top, and $S_j$'s jet is
built from this $\Delta t$-series-valued seed, so every entry of $\mathcal{O}$
becomes a power series in the gaps, $\sum_m R_m\,\Delta t^m$. A constant
identifiability direction $v$ is unobservable for *generic* timing exactly when
$(\sum_m R_m \Delta t^m)v = 0$ for generic $\Delta t$, i.e. when $R_m v = 0$ for
every $m$ (a polynomial vanishing at a generic point). So the $\Delta t$
coefficients are simply stacked as extra rows and reduced over $\mathrm{GF}(p)$ by
the same machinery: nothing symbolic enters the linear algebra, $\Delta t$ only
labels which coefficient layer a row came from.

The gap order is raised, alongside the Lie order, until the rank stops growing
(`info$gapOrderUsed`); truncating early drops rows, which only enlarges
$\ker\mathcal{O}$, so the rank is a **sound conservative lower bound** at every
order and the **exact generic-timing rank** at saturation, just as for the Lie
order. The transcendence is confined to the analytic flow; over the field it is an
honest finite computation (the gap order saturates well below $p$). This recovers
the identifiability of parameters that enter only through a transient between
events: an inhibitor pre-incubation applied at equilibrium, whose relaxed state at
the stimulus is propagated exactly across the gap, identifies the inhibition
strength that the same protocol with the inhibitor collapsed onto $t_0$ cannot.
A `replace` / `add` / `multiply` event on a dynamic state is applied exactly to the
propagated state. (A dose on a conserved-moiety pivot eliminated by `reduceCQ` is
not a state and is not seen; keep that species to dose it.)

## Verifying the rank: the saturation guard {#sec-verify}

The rank verdict rests on one heuristic decision: the Lie order $N$ (and, across
event gaps, the gap order $M$) is raised only until the rank stops growing for a
few consecutive orders, and that plateau is taken as saturation. A finite order
always suffices, because the rank cannot grow beyond $\dim z$; but the automatic
stop, triggered by a handful of flat orders, can be reached *before* that true
saturation. A rank can plateau and then climb again: a quantity that reaches the
output only through a high-order derivative -- a free Hill exponent is the typical
case, whose contribution enters through the recast term $\partial E/\partial n =
E\,L$ (Section \ref{sec-hill}) only once enough time derivatives have accumulated
-- leaves the rank flat for several orders before it lifts. Stopping on that false
plateau drops rows from $\mathcal{O}$, which only enlarges $\ker\mathcal{O}$, so
the failure mode is a silent *over*-reporting: a direction that one more Lie order
would have removed is reported as non-identifiable.

`verify = TRUE`, the default for `"observability"`, closes exactly this gap. It
re-uses the finite-field kernel already built for the analysis and the base point
whose modular steady state is cached, and simply extends the Lie order past the
point where the plateau rule stopped, by `DMOD_SYM_VERIFY_MARGIN` further orders
(default six). Only the Lie jet grows; the expensive $f_{\mathrm{ss}} = 0$ solve is
not repeated, so the check is a handful of matrix ranks at one point, not a second
analysis. If the rank is still climbing past the reported value the plateau was
premature: the guard FAILS and warns, and the directions should be recomputed at a
higher order. If it holds, saturation is confirmed. By the Schwartz--Zippel
argument of Section \ref{sec-finitefield} the modular rank at the point equals the
generic rank with probability one, so a single well-chosen point certifies the
plateau.

The guard is complementary to the per-direction certification of
Section \ref{sec-closedform}, not a repeat of it. That certification re-evaluates
each reconstructed closed form at a fresh prime and rejects a wrong functional
*form*; the scalings are exact integer kernels and need no check. Neither addresses
the *number* of directions, which is fixed by where the Lie order stopped. The
saturation guard is precisely the check on that stopping rule, so the three
together cover the count (guard), the exact directions (integer kernel) and the
reconstructed directions (fresh-prime certification).

`summary()` prints the guard in the computation block, for instance

```
saturation guard: PASSED (rank 233 stable through Lie order 18
  (6 orders beyond the reported saturation))
```

The rank quoted there is *not* the identifiability rank of the header line, and the
two count different things. The header's `rank r / dim d` is the rank after
projecting onto the $d$ unknowns $z$, so its co-rank $d - r$ is the number of
reported directions and $r \le d$ always. The guard's rank is the rank of the raw
observability kernel $\mathcal{O}$ itself, whose columns span the full jet: across
$K$ stacked conditions each carries its own independent copy of the state
coordinates (Section \ref{sec-conditions}), so the kernel has far more columns than
there are shared parameters, and its rank can be many times $d$. Its absolute value
names nothing the modeller reads off; what certifies the verdict is only that it
does *not* change as the Lie order is pushed further. An independent cross-check is
available for small models through `symEngine = "symbolic"`, which builds and
reduces $\mathcal{O}$ over the exact rational-function field with no finite fields
and no recast, carrying the Lie order to the exact saturation bound so that no guard
is needed.

# Supplying the model {#sec-supply}

`symmetryDetection()` is purely symbolic and needs no compiled model: it
takes the equations directly. The principal arguments are

\begin{center}
\begin{tabular}{l p{0.66\textwidth}}
\toprule
argument & meaning \\
\midrule
\texttt{f} & the right-hand sides: an \texttt{eqnlist}, an \texttt{eqnvec}, or a named character vector \\
\texttt{g} & the observation functions, as an \texttt{eqnvec} or named character vector \\
\texttt{trafo} & an optional parameter transformation (parameter-named entries) and initial conditions (state-named entries) \\
\texttt{method} & \texttt{"observability"} (default), \texttt{"polynomial"}, \texttt{"scaling"} or \texttt{"translation"} \\
\bottomrule
\end{tabular}
\end{center}

Anything `as.eqnvec` accepts is a valid `f`; with an `eqnlist` the conserved
moieties are reduced first (`reduceCQ`). Known constants are declared through
`fixed` and so dropped from the unknowns $z$; externally driven inputs through
`forcings`; several experimental conditions through `events` and `conditions`
(Section \ref{sec-conditions}); and the pre-stimulus steady state through a steady-state `trafo`
or `equilibrate` (below). Unless `verify = FALSE`, the reported rank is then
re-checked by the saturation guard of Section \ref{sec-verify}. The two
constructions that deserve more comment are the parameter transformation and the
steady-state constraint.

## Parameter transformations {#sec-trafo}

`trafo` is an [eqnvec] mapping each inner symbol to an expression in the outer
parameters, $\theta^{\mathrm{in}} = \phi(\theta^{\mathrm{out}})$, substituted
symbolically into $f$ and $g$. Three uses occur constantly:

* reparametrisation, for example a rate written through its half-life;
* fixing, $k = 1$, which deletes $k$ from the unknowns (the same effect as
  naming it in `fixed`);
* lumping or a modelling assumption, $k_{\mathrm{deg},m} =
  k_{\mathrm{deg},p} = k_{\mathrm{deg}}$, which ties two inner rates to one
  outer rate.

A state-named `trafo` entry is that state's initial condition rather than a
substitution, so a solved steady state from [steadyStates] can be passed whole.
Whether a reparametrisation can change the reported non-identifiabilities is settled by a
general fact: structural identifiability is invariant under an invertible, smooth
reparametrisation. Such a change of coordinates neither creates nor removes a
non-identifiability, it only changes its form; a scaling in the natural
parameters reappears as a translation in the logarithmic parameters, which is why
the examples below, analysed in natural parameters, report their generators as
scalings. Both engines require rationality, so a `trafo` that makes $f$ or $g$
non-rational, such as $k = \exp(\log k)$, is inadmissible: the flow condition
\eqref{eq:flow-cond} of `"polynomial"` contains $f$ itself, and the finite-field
construction of `"observability"` is built from the field operations only. A
log-parametrised model is therefore analysed in its natural rational parameters,
with identical conclusions.

## Steady-state constraints

Before the experiment a model often rests at a steady state, so the initial
state $x_0$ is not free but solves
\begin{equation}
f_{\mathrm{ss}}(x_0, \theta) = 0,
\label{eq:ssfield}
\end{equation}
the resting right-hand side with all forcings held at zero and no events. This
binds $x_0$ to $\theta$ and can only make the model more identifiable (the role of
the initial conditions in identifiability is classical; Saccomani et al., 2003).
The `"polynomial"` method imposes the binding as one more commutation condition
$[F_{\mathrm{ss}}, X] = 0$ alongside $[F, X] = 0$ and $X(g) = 0$; being an extra
constraint it can only shrink the symmetry set. The `"observability"` method
binds \eqref{eq:ssfield} through the initial condition, and there are two ways to
supply it.

When the steady state has a closed form, pass it through state-named `trafo`
entries: `trafo = steadyStates(reactions)` seeds each state with its rational
expression $x_0(\theta)$, so the order-zero output gradient already carries
$\partial x_0/\partial\theta$. When no closed form exists, set `equilibrate =
TRUE`: the states are then a numeric point on \eqref{eq:ssfield}, recomputed at
each evaluation point and prime. Four facts make this exact and cheap.

*States only, never parameters.* The resting system is solved for the states
alone, treating the rates as fixed symbolic coefficients. Solving instead for a
rate, as a generic symbolic solver may, would express a globally shared rate
through condition-specific quantities and make it falsely look condition-specific;
restricting elimination to the states keeps every global rate global.

*Linear elimination, then a Groebner basis only on the coupled core.* Most states
occur to first degree in some balance with a constant (rate) coefficient and are
solved and substituted directly, a Gaussian elimination over the states that
collapses production and turnover species and leaves only the genuinely nonlinear
core (here the receptor dimerisation and the SMAD complex). That core is a
polynomial system; over $\mathrm{GF}(p)$, with the rates fixed to the current
numeric point, a Groebner basis *triangulates* it into one univariate equation
plus back-substitution. A lexicographic basis triangulates directly but is
ruinously slow on a coupled core, so a cheap grevlex basis is computed and
converted to the lexicographic one by FGLM, which is fast for finitely many
solutions.

*Selecting the interior, consistent point.* The system has several solutions,
including degenerate ones where a species vanishes (saturation by $w\prod_i x_i =
1$, the Rabinowitsch trick, removes those), and the back-substitution must pick a
single consistent branch: each non-zero root is tried and the assembled point is
accepted only when every core equation vanishes. A point chosen blindly per
variable can otherwise mix branches and stall. The rank is constant on the
interior component away from a measure-zero subset, so any generic interior point
gives the generic rank, exactly as for the dynamic matrix of Section \ref{sec-finitefield}.

*Generic-point retries.* When the resting state is algebraic of degree $d$ in the
rates, its coordinates lie in $\mathrm{GF}(p)$ only at the points and primes where
its minimal polynomial splits, roughly a constant fraction of them; the
certification therefore draws successive generic points until one admits a point
over $\mathrm{GF}(p)$ for every condition simultaneously. The parameter
sensitivities come from the implicit function theorem,
$J_x\,(\partial x_0/\partial\theta) = -J_\theta$, solved over $\mathrm{GF}(p)$ at
that point; the closed-form reconstruction and Chinese-remainder lift of
Section \ref{sec-closedform} are unchanged.

The steady state is event-free; the experimental perturbations are applied on top
of it. A switch or held input is a `forcing`, zero at rest and set to its
per-condition value in the post-stimulus dynamics; a dose is a `t0` event,
composed onto the resting initial condition with its parameter sensitivities so
the local jet starts from the stimulated state. State initial conditions given in
`trafo` are ignored under `equilibrate`, which solves them from
\eqref{eq:ssfield}.

## Free Hill and power exponents {#sec-hill}

A rate with a free exponent -- a Hill term $\mathrm{base}^{\,n}/(K^{\,n} +
\mathrm{base}^{\,n})$ or a power $\mathrm{base}^{\,n}$ with $n$ a parameter -- is
not rational, so it would normally be rejected. It is handled instead by a recast,
which needs no `equilibrate` and is an instance of recasting nonlinear systems in
power-law form (Savageau and Voit, 1987). The term $\mathrm{base}^{\,n}$ becomes a
new coordinate $E = \mathrm{base}^{\,n}$ with a companion $L = \log(\mathrm{base})$,
carrying the derivations
\begin{equation}
\begin{aligned}
\dot E &= n\,E\,\frac{\dot{\mathrm{base}}}{\mathrm{base}}, &\quad
\dot L &= \frac{\dot{\mathrm{base}}}{\mathrm{base}}, &\quad
\frac{\partial L}{\partial\,\mathrm{base}} &= \frac{1}{\mathrm{base}}, \\[4pt]
\frac{\partial E}{\partial\,\mathrm{base}} &= n\,\frac{E}{\mathrm{base}}, &\quad
\frac{\partial E}{\partial n} &= E\,L .
\end{aligned}
\end{equation}
With $\mathrm{base}^{\,n}\to E$ the field $f$ is rational and the exponent $n$
enters the observability matrix through $\partial E/\partial n = E\,L$, so it is
assessed like any other parameter. The logarithm is never evaluated:
$\mathrm{base}$, $\log(\mathrm{base})$ and $\mathrm{base}^{\,n}$ are algebraically
independent, so giving $L$ an independent generic residue yields the structural
rank by the same Schwartz-Zippel argument as Section \ref{sec-finitefield}, and the
verification gate of Section \ref{sec-closedform} certifies the verdict at a fresh
prime.

Where the coordinates $E$ and $L$ come from depends on whether the model is
equilibrated. *Without* `equilibrate` the base carries a free initial value, so
$E$ and $L$ are two further free-initial-value coordinates, tied to
$(\mathrm{base}, n)$ by $E = \mathrm{base}^{\,n}$, and the transient itself resolves
the exponent. *With* `equilibrate` the base is a solved steady-state coordinate;
\eqref{eq:ssfield} is then solved with $E$ held at a generic value, an isolated
low-degree system, and the chain rule of $E$ folds into the steady-state Jacobian
so the implicit-function sensitivities $\partial x_0/\partial\theta$ pick up $n$.
As a first, minimal case, a self-inhibited producer with a free Hill exponent,
observed on an unknown scale and analysed *without* a steady state, has its
exponent identified and returns a single weighted scaling:

```{r hill-transient, eval = haveSympy}
f <- eqnvec(p = "kpr / (1 + kinh*p^nhill) - dp*p")     # free Hill exponent nhill
res <- symmetryDetection(f, eqnvec(y = "s*p"), method = "observability",
                         reconstruct = TRUE, reduceCQ = FALSE)
c(rank = res$rank, dim = res$dim,
  nhill_identifiable = !("nhill" %in% res$symmetries[[1]]$support))
res$symmetries[[1]]$generator
```

Rank $5$ of $6$: the Hill exponent `nhill` is identifiable, and the lone direction
is the weighted scaling $\xi_{kinh} = nhill\cdot kinh$, $\xi_{kpr} = -kpr$,
$\xi_p = -p$, $\xi_s = s$ -- rational in `nhill`, with no $\log$ term, since here
the exponent is pinned rather than freed. The recast has done its work with the
base left as an ordinary free-initial-value coordinate.

With `reconstruct`, $E$ and $L$ are sampled as coordinates of the reconstruction
alongside the parameters and back-substituted $E\to\mathrm{base}^{\,n}$,
$L\to\log(\mathrm{base})$, so a direction that genuinely involves the exponent
carries an explicit $\log$ factor, as the example below exhibits. A
minimal self-inhibition makes this concrete: a feedback species `FB`, produced at
`k_pr_FB` and cleared at `d_FB`, throttles the production of `x` through the Hill
term $K^{\,n}/(K^{\,n}+\mathrm{FB}^{\,n})$ with a free exponent `n`, and only a
scaled `x` is measured.

```{r hill-free, eval = haveSympy}
hill <- eqnlist() |>
  addReaction("0",  "FB", "k_pr_FB")                     |>
  addReaction("FB", "0",  "d_FB * FB")                   |>
  addReaction("0",  "x",  "k_pr_x * K^n / (K^n + FB^n)") |>
  addReaction("x",  "0",  "d_x * x")

# the modular engine assesses the free exponent n through the recast, like any
# parameter: n and the Michaelis constant K come out non-identifiable (dim 7)
res <- symmetryDetection(hill, eqnvec(xobs = "scale * x"),
                         method = "observability", equilibrate = TRUE,
                         reduceCQ = FALSE)
c(identifiable = res$identifiable, dim = res$dim,
  n_confounded = "n" %in% unlist(lapply(res$symmetries, function(d) d$support)))

# the explicit direction that frees n, exhibited with the symbolic engine (which
# carries base^n as an atom, so a free exponent is native). It needs the steady
# state supplied, as it does not equilibrate: n trades against K with a log factor.
ss <- eqnvec(FB = "k_pr_FB/d_FB",
             x  = "k_pr_x*K^n/(d_x*(K^n + (k_pr_FB/d_FB)^n))")
sres <- symmetryDetection(hill, eqnvec(xobs = "scale * x"),
                          method = "observability", symEngine = "symbolic", trafo = ss)
Filter(function(d) "n" %in% d$support, sres$symmetries)[[1]]$generator
```

The exponent direction has $\xi_n = 1$ and
\begin{equation*}
\xi_K = \frac{K\log(k_{pr\_FB}/d_{FB}) - K\log K}{n}
      = \frac{K}{n}\,\log\!\frac{k_{pr\_FB}/d_{FB}}{K},
\end{equation*}
so the free exponent $n$ and the Michaelis constant $K$ cannot be told apart (the
resting feedback level $k_{pr\_FB}/d_{FB}$ is the other base), and the
$\log(\mathrm{base})$ factor predicted above appears explicitly. A companion case is a feedback whose free exponent is itself
identifiable while its inhibition strength confounds with the regulator's synthesis
rate. That direction is a scaling weighted by the exponent, $\xi_{k_{fb}} = -n\,k_{fb}$
traded against $\xi_{k_{pf}} = k_{pf}$ (the invariant is $k_{fb}\,k_{pf}^{\,n}$),
recovered in closed form from its minimal support in log coordinates rather than by
sampling the recast exponent. Entered linearly, as several feedbacks are in the
SMAD model of Section \ref{sec-examples}, the same confounder is an ordinary integer
scaling; entered through a Hill term, as the receptor feedbacks there are, it keeps
the exponent in its weight.

A feedback produced from the observed species makes this concrete. Only `x` is
measured; a dose renders the exponent `n` identifiable, and the inhibition strength
`k_fb` then confounds with the feedback synthesis `k_pf`:

```{r hill-weighted, eval = haveSympy}
loop <- eqnvec(x  = "k_pr/(1 + k_fb*FB^n) - d_x*x + k_stim*u",
               FB = "k_pf*x - d_FB*FB", u = "0")
res <- symmetryDetection(loop, eqnvec(xobs = "s*x"), method = "observability",
  equilibrate = TRUE, forcings = "u",
  events = addEvent(eventlist(), var = "u", time = 0, value = "dose",
                    method = "replace"),
  conditions = data.frame(dose = 1, row.names = "c1"),
  reconstruct = TRUE, reduceCQ = FALSE)
Filter(function(d) setequal(d$support, c("k_fb", "k_pf")),
       res$symmetries)[[1]]$generator
```

The confounder comes back as $\xi_{k_{fb}} = -n\,k_{fb}$ against
$\xi_{k_{pf}} = k_{pf}$ (the same direction as above up to scale, normalised on
`k_pf` here), the weighted scaling with the free exponent `n` carried in the weight
and no $\log$ term, since here the exponent is identified rather than freed.

### Bases without a turnover term

The recast above keeps $\mathrm{base}$ a solved state and holds $E$ generic, which
relies on $\mathrm{base}$ appearing with a linear turnover term in its own balance
(the $-d_p\,p$ above). When it does not, as in a pure power $k_{pr} -
d_p\,\mathrm{base}^{\,n} + k_{in}u$, the recast balance is linear in $E$ and pins
$E$ rationally, while $\mathrm{base} = \big((k_{pr}+k_{in}u)/d_p\big)^{1/n}$ is a
fractional root. The roles then invert: $E$ is the solved coordinate and
$\mathrm{base}$, $L = \log(\mathrm{base})$ are held generic, with duals from the
inverse chain rule
\begin{equation}
\begin{aligned}
\frac{\partial\,\mathrm{base}}{\partial\theta} &=
  \frac{\mathrm{base}}{n\,E}\,\frac{\partial E}{\partial\theta}
  - [\theta = n]\,\frac{\mathrm{base}\,L}{n}, \\
\frac{\partial L}{\partial\theta} &=
  \frac{1}{n\,E}\,\frac{\partial E}{\partial\theta}
  - [\theta = n]\,\frac{L}{n},
\end{aligned}
\end{equation}
read off $\mathrm{base} = E^{1/n}$, $L = \tfrac{1}{n}\log E$ and $\log E = n\,L$. The
same algebraic independence holds with $E$ now solved, so the structural rank
follows as before. A $\log$ survives in a closed-form direction only when a symmetry
moves a state along the free-exponent power; the pure production $\dot x = k_{pr} -
d_p\,x^{\,q} + k_{in}u$ with $y = s\,x$ instead has the rational scaling direction
$\xi_{d_p} = (1-q)\,d_p$, $\xi_s = -s$, restoring rationality with no $\log$. A base
carrying two independent free-exponent powers and no turnover term has no rational
handle and is reported as non-rational.

# Examples {#sec-examples}

```{r lib, eval = haveSympy, message = FALSE}
library(dMod)
```

## A gene-expression cascade at steady state

A stimulus `TF` drives transcription, the mRNA is translated into protein, both
species degrade (here at a shared rate, a common modelling assumption), and only
a scaled protein readout is measured, as for a fluorescence or immunoblot signal
of unknown gain. Before stimulation the cell rests at the basal steady state, so
the initial state is not free.

```{r genex-model, eval = haveSympy}
eq <- eqnlist() |>
  addReaction("",     "mRNA", "ksyn * TF")   |>
  addReaction("mRNA", "",     "kdeg * mRNA") |>
  addReaction("",     "Prot", "ktrl * mRNA") |>
  addReaction("Prot", "",     "kdeg * Prot")

g <- eqnvec(yProt = "scale * Prot")          # readout of unknown gain
```

The whole analysis is symbolic; no model is compiled. The pre-equilibration is
imposed with `equilibrate = TRUE`, which puts the states on the resting manifold
$f_{\mathrm{ss}} = 0$ by the construction of the steady-state constraint above, so
no steady state need be solved by hand.

```{r genex-oim, eval = haveSympy}
res <- symmetryDetection(eq, g, method = "observability",
                         equilibrate = TRUE, reconstruct = TRUE)
summary(res)
```

At rest the protein sits at $\mathrm{scale}\cdot TF\,ksyn\,ktrl/kdeg^2$, so only
that one product is observed: the matrix has rank $1$ of $5$, leaving four free
directions. All four are exact integer scalings, peeled from the stacked kernel
and printed as their toric weights. Three trade the numerator factors of that
product — $ksyn$ against $ktrl$, $TF$ against $ksyn$, and the gain $scale$ against
the translation level. The fourth couples $TF$, $ktrl$ and $kdeg$ with equal
weight: it is induced by the steady-state constraint, where the degradation enters
squared, so scaling $kdeg$ up is offset by scaling the input and the translation
rate up (the $kdeg^2$ in the denominator absorbs the two numerator factors).
Handing the steady state in explicitly with `trafo = steadyStates(eq)` gives the
identical verdict; `equilibrate` reproduces it, recomputing the modular steady
state and its sensitivities at each point and prime. The summary's computation
block ends with the saturation guard of Section \ref{sec-verify}, confirming that
the rank has stopped growing so the four directions are neither too many nor too
few.

## Acting on a result: reparametrisation

`symmetryDetection()` does not change the model; it names the combinations the
data constrain, and the modeller removes the redundancy. A non-identifiable
direction almost always means that a product or ratio of parameters is
identifiable while the individual factors are not, so the remedy is to make that
combination a single parameter and fix the redundant factor.

When a factor is independently known, fix it. The gene cascade sees only
`scale * Prot`, confounding the readout gain `scale` with the absolute protein
level; an absolute calibration is imposed with `fixed`, which removes that
direction:

```{r genex-fix, eval = haveSympy}
res2 <- symmetryDetection(eq, g, method = "observability",
                          equilibrate = TRUE, fixed = "scale",
                          reconstruct = TRUE)
c(rank = res2$rank, dim = res2$dim)
```

More often no factor is separately known and the product itself is the
meaningful quantity. If a flux enters as `k1 * k2 * x`, only the product
`k1 * k2` is identifiable, never `k1` and `k2` apart. Carry it as one parameter
by fixing the redundant factor to one in a `trafo`, letting `k = k1 * k2` absorb
the product:

```{r reparam, eval = haveSympy}
f <- eqnvec(x = "k1*k2 - d*x")            # only the product k1*k2 is identifiable
symmetryDetection(f, eqnvec(y = "x"), method = "observability")$identifiable

symmetryDetection(f, eqnvec(y = "x"), method = "observability",
                  trafo = eqnvec(k1 = "1", k2 = "k"))$identifiable   # now TRUE
```

The model is now estimated in `k`. The same recipe resolves any scaling
direction $\sum_i c_i\, z_i\, \partial_{z_i}$: fix all but one of the involved
coordinates (to one, or to a known value) and let the survivor carry the product. A reparametrisation is
admissible as long as it keeps the right-hand sides rational (Section \ref{sec-trafo});
`k = exp(log k)` is not.

The cascade observed dynamically from a free initial state, rather than at rest,
gives the rate curve worked out below under *Two rates on a curve*.

## Closed-form directions and a known input

A non-identifiable direction is a vector in parameter space, but its components
may be rational functions of a known input: when two parameters enter only
through a combination weighted by that input, the trade-off between them carries
its weight. In `A' = k1 + u*k2 - kdeg*A` observed through `y = A`, the influx
`k1 + u*k2` is lumped, so `k1` and `k2` cannot be separated; declaring the known
constant `u` with `fixed` keeps it out of the unknowns yet lets it appear as a
coefficient, and the direction is returned in closed form.

```{r toy-input, eval = haveSympy}
f   <- eqnvec(A = "k1 + u*k2 - kdeg*A")
toy <- symmetryDetection(f, eqnvec(y = "A"), method = "observability",
                         reconstruct = TRUE, fixed = "u")
Filter(function(d) d$type != "scaling", toy$symmetries)
```

The direction is the constant `k1 : u`, `k2 : -1` (the input-weighted slope
\eqref{eq:toy-null}, reported up to sign): a unit change in `k2` is compensated by
`-u` in `k1`. The constant `u` never enters
$z$; it appears only as the coefficient of the trade-off. When such an input
takes several known values across conditions, stacking those conditions
separates `k1` and `k2` outright, and `u`, now a constant of each condition,
drops out of every reported direction.

## A pre-equilibrated model with a dose

Most experiments do not start from an arbitrary state: the system is left to
reach a resting steady state, a stimulus is applied, and the relaxation is
recorded. The steady state is then not a free initial condition but is fixed by
the dynamics, and that constraint is part of the model. A single species makes
the whole calculation explicit. Let

\begin{equation}
\dot x \;=\; b - a\,x, \qquad y \;=\; s\,x,
\label{eq:ss-model}
\end{equation}

with basal production $b$, first-order turnover $a$ and a readout scale $s$.
Before stimulation the system rests where $\dot x = 0$, the unique steady state

\begin{equation}
x_\ast \;=\; b/a .
\label{eq:ss-fixed}
\end{equation}

At $t = 0$ a dose sets the level to $D$ (a `replace` event), and the species
relaxes back towards $x_\ast$,

\begin{equation}
x(t) \;=\; \frac{b}{a} + \Bigl(D - \frac{b}{a}\Bigr) e^{-a t},
\qquad y(t) = s\,x(t).
\label{eq:ss-traj}
\end{equation}

**The output coefficients.** The Taylor coefficients of $y$ at $t = 0$ follow
from \eqref{eq:ss-traj} by repeated differentiation and use of $\dot x = b - ax$:

\begin{equation}
y(0) = s D, \qquad
\dot y(0) = s\,(b - a D), \qquad
\ddot y(0) = s\,a\,(a D - b),
\label{eq:ss-taylor}
\end{equation}

each a multiple of the single relaxation mode, with the rate $a$ fixing the
geometric ratio between successive coefficients. Differentiating
\eqref{eq:ss-taylor} in the unknowns $z = (a, b, s, D)$ gives the OIM rows

\begin{equation}
\mathcal{O} \;=\;
\begin{pmatrix}
0          & 0    & D          & s    \\
-sD        & s    & b - aD     & -sa  \\
2saD - sb  & -sa  & a^2 D - ab & sa^2
\end{pmatrix},
\label{eq:ss-oim}
\end{equation}

the columns ordered $(a, b, s, D)$. The rate $a$ is identifiable from the ratio
of successive coefficients; beyond it the data carry only the two amplitude
products $sD$ (the height of the jump, from $y(0)$) and $sb$ (through the
asymptote $s x_\ast = sb/a$, with $a$ known).

**Dose unknown.** With $D$ among the unknowns, $\mathcal{O}$ has rank $3$ and a
one-dimensional null space, spanned by

\begin{equation}
\xi:\quad a \mapsto 0,\quad b \mapsto -b,\quad s \mapsto s,\quad D \mapsto -D,
\label{eq:ss-null}
\end{equation}

scaling the readout up while scaling every amount down. One checks
$\mathcal{O}\,\xi = 0$ row by row: row $0$ is $Ds - sD = 0$, row $1$ is
$-sb + (b - aD)s + saD = 0$, row $2$ is $sab + (a^2 D - ab)s - sa^2 D = 0$. The
products $sb$ and $sD$ are fixed along $\xi$, but $s$ alone is not: the readout
scale cannot be told from the absolute amounts. This is the same calibration
confounder as in the canonical example, now riding on a pre-equilibrated,
dose-perturbed trajectory.

**Dose known.** When the dose $D$ is set by the experiment, $sD$ pins $s$, then
$sb$ gives $b$, and with $a$ already fixed the model is structurally
identifiable. The pre-equilibration is what makes this clean: were the initial
value a free unknown rather than $x_\ast = b/a$, a second non-identifiability
would appear (the resting level against the relaxation amplitude); tying the
start to the steady state removes that freedom.

**Verifying with the method.** The dose is a `replace` event; a condition grid
carries its value. A known dose gives a structurally identifiable model, an
unknown dose magnitude returns exactly the confounder \eqref{eq:ss-null}.

```{r ss-dose, eval = haveSympy}
f <- eqnvec(x = "b - a*x")     # turnover; steady state b/a
g <- eqnvec(y = "s*x")         # scaled readout
dose <- addEvent(eventlist(), var = "x", time = 0, value = "dose",
                 method = "replace")

# a known dose: the relaxation identifies the rate, the scale and the production
symmetryDetection(f, g, method = "observability", events = dose,
                  conditions = data.frame(dose = 2,
                                          row.names = "stim"))$identifiable

# an unknown dose magnitude: the readout scale confounds with the amounts
res <- symmetryDetection(f, g, method = "observability", reconstruct = TRUE,
                         events = dose,
                         conditions = data.frame(row.names = "stim"))
res$symmetries[[1]]$generator
```

The reported direction is the exact integer scaling $b:1,\ \mathrm{dose}:1,\ s:-1$
(printed as `b : b`, `dose : dose`, `s : -s`): the same direction as
\eqref{eq:ss-null} up to sign, scaling the amounts up while scaling the readout
gain down. It is peeled from the integer kernel, so no rational reconstruction is
needed. A genuinely rational, non-scaling direction — the kind the multi-point
reconstruction of Section \ref{sec-closedform} lifts through the
Chinese-remaindering step worked through there for $-\tfrac{1}{3}$ — appears in
the partially observed cascade below. Pre-equilibrated states, a dose set by an
event, and several conditions stacked is the standard pattern, and it scales
unchanged to large signalling models.

The same resting state can be imposed without writing it down. Here $f = b - ax$
rests at $x^* = b/a$, but in general the steady state need not be a rational
expression. Setting `equilibrate = TRUE` lets the method find an interior point of
$f_{\mathrm{ss}} = 0$ numerically (per prime, over a finite field) and seed it
with its parameter sensitivities, so no closed form is required and the verdict
matches the explicit `trafo = eqnvec(x = "b/a")`:

```{r ss-constraint, eval = haveSympy}
# rank 1 of 3: only the resting product s*b/a is seen, exactly as with the
# explicit steady state trafo = eqnvec(x = "b/a")
res <- symmetryDetection(f, g, method = "observability",
                         equilibrate = TRUE, reconstruct = TRUE)
c(identifiable = res$identifiable, rank = res$rank, dim = res$dim)

# a genuinely irrational steady state x* = sqrt(b/a): no rational form exists,
# yet the modular interior point still delivers the verdict
symmetryDetection(eqnvec(x = "b - a*x^2"), g, method = "observability",
                  equilibrate = TRUE)$rank
```

## A transient channel made identifiable by exact propagation

The exact gap propagation of Section \ref{sec-segments} identifies parameters that
enter only through a transient between events. A species $x$ is produced under a
feedback inhibition of strength $k_{\mathrm{inh}}$, switched on by an inhibitor
input, and read out on an unknown scale; a stimulus follows later. The inhibition
strength affects only the inhibitor-relaxed state at the stimulus onset, not any
directly observed quantity. Applied before the stimulus and propagated exactly across the gap between the two
event times, $k_{\mathrm{inh}}$ becomes identifiable; with the inhibitor and
stimulus at the same time (no gap), it stays confounded. The analysis starts at the
earliest event time on its own, so the two cases differ only in the inhibitor's
event time.

```{r segments, eval = haveSympy}
f <- eqnvec(x = "kpr/(1 + kinh*inh) - kdeg*x + kstim*stim",
            inh = "0", stim = "0")
g <- eqnvec(y = "s*x")

# inhibitor and stimulus at the same time (no gap): kinh not identifiable
ev0 <- eventlist() |>
  addEvent(var = "inh",  time = 0, value = "1", method = "replace") |>
  addEvent(var = "stim", time = 0, value = "1", method = "replace")
r0 <- symmetryDetection(f, g, method = "observability", equilibrate = TRUE,
                        events = ev0, forcings = c("inh", "stim"))
"kinh" %in% unlist(lapply(r0$symmetries, function(d) d$support))   # TRUE

# inhibitor applied 30 units before the stimulus: the gap is propagated exactly
ev30 <- eventlist() |>
  addEvent(var = "inh",  time = -30, value = "1", method = "replace") |>
  addEvent(var = "stim", time = 0,   value = "1", method = "replace")
r30 <- symmetryDetection(f, g, method = "observability", equilibrate = TRUE,
                         events = ev30, forcings = c("inh", "stim"))
c(gapOrder = r30$info$gapOrderUsed,
  kinhIdentifiable = !("kinh" %in% unlist(lapply(r30$symmetries,
                                                 function(d) d$support))))
```

The second run reports a non-zero `info$gapOrderUsed` (the gap series carries
$k_{\mathrm{inh}}$) and a higher rank than the first.

## An enzyme assay and its hidden symmetries

The examples so far were designed so the symmetry could be read off the
equations. A more realistic case is the classical enzyme-kinetics assay, where
the non-identifiabilities are genuinely not visible by inspection yet can still
be derived by hand and then confirmed by the method. A substrate is turned over
by an enzyme under the quasi-steady-state (Michaelis--Menten) law, and the
substrate is read out by some reporter of unknown gain $s$:
\begin{equation}
\dot S = -\,\frac{k_{\mathrm{cat}}\,E_{\mathrm{tot}}\,S}{K_m + S},
\qquad y = s\,S.
\label{eq:mm}
\end{equation}
The unknowns are the initial substrate $S_0$, the total enzyme
$E_{\mathrm{tot}}$, the Michaelis constant $K_m$, the readout gain $s$ and the
turnover number $k_{\mathrm{cat}}$, so $\dim z = 5$.

Two substitutions leave the entire observed trajectory $y(t)$, and therefore
every time derivative that the observability matrix collects, unchanged.

*The turnover--enzyme confounder.* The pair $(k_{\mathrm{cat}}, E_{\mathrm{tot}})$
enters \eqref{eq:mm} only through the product $V_{\max} = k_{\mathrm{cat}}\,
E_{\mathrm{tot}}$. Hence
\begin{equation}
k_{\mathrm{cat}} \to \lambda\,k_{\mathrm{cat}}, \qquad
E_{\mathrm{tot}} \to \lambda^{-1} E_{\mathrm{tot}}
\end{equation}
fixes $V_{\max}$, so $\dot S$, $S(t)$ and $y$ are untouched: only $V_{\max}$ is
identifiable, never the turnover number and the enzyme amount separately. The
generator is $k_{\mathrm{cat}}\partial_{k_{\mathrm{cat}}} -
E_{\mathrm{tot}}\partial_{E_{\mathrm{tot}}}$, the integer toric weights
$(k_{\mathrm{cat}}\!:1,\; E_{\mathrm{tot}}\!:-1)$.

*The readout-units confounder.* We never see $S$ in molar units, only $y = sS$.
The right-hand side of \eqref{eq:mm} is jointly homogeneous of degree one in the
amounts $(S, K_m, E_{\mathrm{tot}})$: replacing each by $\mu\cdot(\,\cdot\,)$
multiplies $\dot S$ by $\mu$, so $\tilde S = \mu S$ solves the rescaled system.
Compensating the gain by $s \to \mu^{-1}s$ keeps $y = sS$ fixed. The generator is
$S\partial_S + K_m\partial_{K_m} + E_{\mathrm{tot}}\partial_{E_{\mathrm{tot}}} -
s\,\partial_s$, the integer toric weights
$(S\!:1,\; E_{\mathrm{tot}}\!:1,\; K_m\!:1,\; s\!:-1)$.

Both substitutions are scalings, so the common invariants, the quantities the
data actually pin down, are $sS$ (the readout itself), $sK_m$ and
$s\,k_{\mathrm{cat}}E_{\mathrm{tot}} = sV_{\max}$: three functions, matching a
rank of three out of five. Both are scalings, so the `"observability"` engine
extracts them from the integer kernel and reports them as exact toric weights:

```{r mm, eval = haveSympy}
f <- eqnvec(S = "-kcat*Etot*S/(Km + S)")   # quasi-steady-state turnover
g <- eqnvec(y = "s*S")                      # readout of unknown gain s

res <- symmetryDetection(f, g, method = "observability", reconstruct = TRUE)
summary(res)
```

The engine returns two integer scalings spanning the same rank-$2$ lattice (each
carries `type = "scaling"`). The printed basis may combine the two hand-derived
confounders — a null space has no canonical basis — but the invariants it fixes
are exactly $sS$, $sK_m$ and $sV_{\max}$, so the verdict is the same. The
dedicated `"scaling"` engine returns the same lattice.

The verdict and the involved symbols come from modular rank computations at
generic points, not from a symbolic determining system, so `"observability"`
also applies to networks with dozens of states and conserved moieties. There the
closed-form reconstruction is attempted only for directions that couple few
unknowns; one that couples too many is returned with its support and
`reconstruct = FALSE`.

## Two rates on a curve

A non-identifiability need not be a straight ray through parameter space. The
plainest example is gene expression observed only at the protein level, an mRNA
$m$ transcribed at rate $k_{tx}$ and degraded at $d_m$, translated into protein
$p$ at rate $k_{tl}$ and degraded at $d_p$:
\begin{equation}
\dot m = k_{tx} - d_m\,m, \qquad
\dot p = k_{tl}\,m - d_p\,p, \qquad y = p.
\label{eq:gene}
\end{equation}
The mRNA is never measured. Then transcription and translation cannot be told
apart: more mRNA poorly translated and less mRNA well translated give the same
protein. Concretely, scaling the hidden mRNA by any $\lambda > 0$,
\begin{equation}
m \to \lambda\,m, \quad m_0 \to \lambda\,m_0, \quad
k_{tx} \to \lambda\,k_{tx}, \quad k_{tl} \to \lambda^{-1} k_{tl},
\label{eq:gene-sym}
\end{equation}
leaves $y = p$ invariant: the first equation becomes
$\dot{(\lambda m)} = \lambda k_{tx} - d_m(\lambda m)$, the transcription term
scaling with $m$, while the protein equation keeps $k_{tl} m = (\lambda^{-1}
k_{tl})(\lambda m)$ unchanged. The degradation rates $d_m, d_p$, set by the two
observed timescales, stay fixed.

Eliminating $\lambda$ from \eqref{eq:gene-sym}, the two rate constants are
confined to the level set
\begin{equation}
k_{tx}\,k_{tl} = \text{const},
\end{equation}
a hyperbola, not a line: any $(k_{tx}, k_{tl})$ on it produces exactly the same
protein trajectory (only their product, which sets the steady state
$p_\infty = k_{tx} k_{tl}/(d_m d_p)$, is identifiable). The orbit is curved, but
its generator is a plain scaling: the integer toric weights
$(k_{tx}\!:1,\; k_{tl}\!:-1,\; m\!:1)$, which is, up to its overall sign, the
single direction the method returns:

```{r gene, eval = haveSympy}
f <- eqnvec(m = "ktx - dm*m", p = "ktl*m - dp*p")   # mRNA hidden, protein seen
res <- symmetryDetection(f, eqnvec(y = "p"), method = "observability",
                         reconstruct = TRUE)
res$rank                                  # 5 of 6: one direction
res$symmetries[[1]]$generator
```

The opposite weights $k_{tx}\!:1$, $k_{tl}\!:-1$ trace the hyperbola
$k_{tx} k_{tl} = \text{const}$, and the $m\!:1$ weight records that the hidden
mRNA scales with $k_{tx}$ along it. Degradation rates and the protein do not
appear: the data fix them, and confine $k_{tx}, k_{tl}$ to the curve but no
further.

## EGF/EGFR feeding a MEK/ERK cascade

The directions so far were single scalings. A slightly larger but still
hand-checkable network shows how to *read* a whole set of directions at once, and
how each one turns back into an explicit transformation $\Phi_\varepsilon$ by the
integration of Section \ref{sec-flow}. Epidermal growth factor binds its receptor
to form an active complex that phosphorylates MEK, and phospho-MEK in turn
phosphorylates ERK; each kinase is returned by a constitutive phosphatase. Only
the two phospho-forms are measured, each on an unknown gain, so the ligand, the
receptor and the unphosphorylated kinase pools are hidden: the model is partially
observed. Every rate is mass-action, so every direction is rational and is
returned in closed form.

```{r egf-model, eval = haveSympy}
reactions <- eqnlist() |>
  addReaction("EGF + EGFR", "EGF_EGFR", "k_bind * EGF * EGFR")   |>
  addReaction("EGF_EGFR", "EGF + EGFR", "k_unbind * EGF_EGFR")   |>
  addReaction("MEK", "pMEK", "k_phos_MEK * EGF_EGFR * MEK")      |>
  addReaction("pMEK", "MEK", "k_dephos_MEK * pMEK")              |>
  addReaction("ERK", "pERK", "k_phos_ERK * pMEK * ERK")          |>
  addReaction("pERK", "ERK", "k_dephos_ERK * pERK")
reactions <- customTotals(reactions, list(
  totalEGF  = "EGF + EGF_EGFR", totalEGFR = "EGFR + EGF_EGFR",
  totalMEK  = "MEK + pMEK",     totalERK  = "ERK + pERK"))

observables <- eqnvec(pMEK_obs = "scale_pMEK * pMEK",
                      pERK_obs = "scale_pERK * pERK")
```

Running with `reduceCQ = FALSE` reports the directions in the species themselves
rather than through the moiety totals, which makes them easier to read:

```{r egf-run, eval = haveSympy}
egf <- symmetryDetection(reactions, observables, method = "observability",
                         reduceCQ = FALSE, reconstruct = TRUE)
summary(egf)
```

Rank $11$ of $15$: four directions, three scalings and one genuinely rational
(non-scaling) confounder. Each is forced by the partial observation. Every
direction is printed as its differential generator $\sum_i \xi_i\, \partial_{z_i}$
with its symmetry class (`scaling`, `polynomial`, ...) in brackets.

* **Receptor-module units**
  $(EGF\!:\!1,\, EGFR\!:\!1,\, EGF\_EGFR\!:\!1,\, k_{bind}\!:\!-1,\, k_{phosMEK}\!:\!-1)$.
  The ligand, the receptor and their complex are never seen directly, so their
  absolute concentration scale is free. Multiplying all three by $\lambda$ scales
  the bimolecular flux $k_{bind}\,EGF\,EGFR$ by $\lambda^2$, undone by
  $k_{bind}\!\to\!k_{bind}/\lambda$; and the activation of MEK,
  $k_{phosMEK}\,EGF\_EGFR\,MEK$, is held fixed by
  $k_{phosMEK}\!\to\!k_{phosMEK}/\lambda$, so the cascade downstream does not move.
* **MEK amount against its readout gain**
  $(MEK\!:\!1,\, pMEK\!:\!1,\, k_{phosERK}\!:\!-1,\, scale_{pMEK}\!:\!-1)$.
  The total MEK is unknown in absolute units: scaling MEK and pMEK by $\lambda$ is
  absorbed by the gain $scale_{pMEK}\!\to\!scale_{pMEK}/\lambda$ (the readout
  $scale_{pMEK}\,pMEK$ is fixed) and by $k_{phosERK}\!\to\!k_{phosERK}/\lambda$
  (the ERK activation $k_{phosERK}\,pMEK\,ERK$ is fixed).
* **ERK amount against its readout gain**
  $(ERK\!:\!1,\, pERK\!:\!1,\, scale_{pERK}\!:\!-1)$. The same calibration
  trade-off at the foot of the cascade; nothing lies downstream of ERK, so no
  rate compensates.
* **The binding equilibrium** (polynomial). Only $EGF$, $EGFR$ and $k_{unbind}$
  move, with the active complex $EGF\_EGFR$ and $k_{bind}$ held fixed. Because the
  cascade reads the receptor module only through that complex, the dissociation
  rate and the split between free ligand and free receptor cannot be told apart:
  they trade while the complex, and therefore everything observed, stays put. This
  is the one non-monomial direction, returned as the rational vector above.

**Reading a direction as a transformation.** A reported direction is the generator
$\xi$; the transformation it stands for is its flow $\Phi_\varepsilon$ from
Section \ref{sec-flow}. For a scaling the integration \eqref{eq:scaling-flow} is
immediate. The receptor-module direction integrates, with $\lambda = e^\varepsilon$,
to
\begin{equation}
\begin{aligned}
EGF &\mapsto \lambda\,EGF, &\quad
EGFR &\mapsto \lambda\,EGFR, &\quad
EGF\_EGFR &\mapsto \lambda\,EGF\_EGFR, \\
k_{bind} &\mapsto \lambda^{-1} k_{bind}, &\quad
k_{phosMEK} &\mapsto \lambda^{-1} k_{phosMEK},
\end{aligned}
\end{equation}
every other coordinate fixed. Substituting into $k_{bind}\,EGF\,EGFR$ and
$k_{phosMEK}\,EGF\_EGFR\,MEK$ returns both fluxes to their values, so the output is
unchanged for every $\lambda > 0$: the reported integer weights are precisely this
power law.

The general direction is the same construction with a non-constant $\xi$. Writing
$C = EGF\_EGFR$ and $D = k_{bind}(EGF - EGFR)$, the flow \eqref{eq:flow} is
\begin{equation}
\frac{d\,EGF}{d\varepsilon} = \frac{-(EGF + C)}{D},\qquad
\frac{d\,EGFR}{d\varepsilon} = \frac{EGFR + C}{D},\qquad
\frac{d\,k_{unbind}}{d\varepsilon} = 1,
\end{equation}
with $C$, $k_{bind}$ and everything downstream constant. It has no elementary
closed form, but integrating it explicitly is unnecessary to see what it does:
along the flow
\begin{equation}
\frac{d}{d\varepsilon}\big(k_{bind}\,EGF\,EGFR\big) = C,\qquad
\frac{d}{d\varepsilon}\big(k_{unbind}\,EGF\_EGFR\big) = C,
\end{equation}
so the binding and unbinding fluxes move in lockstep and their difference, the net
rate of complex formation, is preserved; $EGF\_EGFR$ itself is fixed
($\xi_{EGF\_EGFR} = 0$) while $k_{unbind}$ simply advances by $\varepsilon$. The
orbit re-partitions free ligand and receptor and shifts the dissociation rate
while holding the active complex and its dynamics, hence the entire measured
cascade, invariant: exactly the non-identifiability the support announced. This is
the general pattern. `symmetryDetection()` returns the generator $\xi$ exactly, and
the finite transformation $\Phi_\varepsilon$ follows by integrating it, in closed
form for a scaling and as a well-defined flow otherwise.

## A signalling network at scale

The same analysis runs on a model far larger than a symbolic determining system
can handle, and it exercises every ingredient at once: many conditions, event
gaps, a modularly solved steady state, and free Hill exponents. The example is a
two-receptor TGF$\beta$/SMAD signalling model with $26$ dynamic states and three
conserved SMAD moieties: receptor mRNA transcription under Hill-type
transcriptional feedback (a free exponent on each feedback species), translation
and degradation, ligand binding to each receptor and to the preformed receptor
complex, receptor internalisation, the active signalling complex, SMAD2 and SMAD3
phosphorylation modulated by feedback, the trimeric SMAD2--SMAD3--SMAD4 complex,
and three feedback genes (`FB2`, `FB3`, `FB4`). The reaction list is

```{r smad-model, eval = haveSympy}
addRC <- function(eq, from, to, rate, ...)
  addReaction(eq, from, to, rate, compartment = "Cell", ...)
addRE <- function(eq, from, to, rate, ...)
  addReaction(eq, from, to, rate, compartment = "extraCell", ...)

reactions <- eqnlist() |>
  addRC("", "bool_ActD", "0") |>
  addRC("", "bool_CHX", "0") |>
  addRC("", "bool_MG132", "0") |>
  addRE("", "TGFb", "0") |>
  addRC("", "R1mRNA",  "k_pr_R1mRNA * (1 + k_inh_R1mRNA_FB3 * FB3^nhill_R1) * (1 - bool_ActD)") |>
  addRC("", "R2mRNA",  "k_pr_R2mRNA / (1 + k_inh_R2mRNA_FB4 * FB4^nhill_R2) * (1 - bool_ActD)") |>
  addRC("", "FB2mRNA", "k_pr_FB2mRNA * C3^nhill_FB2mRNA / (Km_FB2mRNA^nhill_FB2mRNA + C3^nhill_FB2mRNA) * (1 - bool_ActD)") |>
  addRC("", "FB3mRNA", "k_pr_FB3mRNA * C3^nhill_FB3mRNA / (Km_FB3mRNA^nhill_FB3mRNA + C3^nhill_FB3mRNA) * (1 - bool_ActD)") |>
  addRC("", "FB4mRNA", "k_pr_FB4mRNA * C3^nhill_FB4mRNA / (Km_FB4mRNA^nhill_FB4mRNA + C3^nhill_FB4mRNA) * (1 - bool_ActD)") |>
  addRC("R1mRNA", "", "k_dg_R1mRNA * R1mRNA") |>
  addRC("R2mRNA", "", "k_dg_R2mRNA * R2mRNA") |>
  addRC("FB2mRNA", "", "k_dg_FB2 * FB2mRNA") |>
  addRC("FB3mRNA", "", "k_dg_FB3 * FB3mRNA") |>
  addRC("FB4mRNA", "", "k_dg_FB4 * FB4mRNA") |>
  addRC("", "R1", "k_pr_R1 * R1mRNA * (1 - bool_CHX)") |>
  addRC("", "R2", "k_pr_R2 * R2mRNA * (1 - bool_CHX)") |>
  addRC("", "FB2", "k_pr_FB2 * FB2mRNA * (1 - bool_CHX)") |>
  addRC("", "FB3", "k_pr_FB3 * FB3mRNA * (1 - bool_CHX)") |>
  addRC("", "FB4", "k_pr_FB4 * FB4mRNA * (1 - bool_CHX)") |>
  addRC("R1", "", "k_dg_R1 * R1 * (1 - bool_MG132)") |>
  addRC("R2", "", "k_dg_R2 * R2 * (1 - bool_MG132)") |>
  addRC("FB2", "", "k_dg_FB2 * FB2 * (1 - bool_MG132)") |>
  addRC("FB3", "", "k_dg_FB3 * FB3 * (1 - bool_MG132)") |>
  addRC("FB4", "", "k_dg_FB4 * FB4 * (1 - bool_MG132)") |>
  addRC("R1 + R2", "R1_R2", "k_act_R1_R2 * R1 * R2") |>
  addRC("R1_R2", "R1 + R2", "k_deact_R1_R2 * R1_R2") |>
  addRC("R1_R2", "", "k_dg_R1_R2 * R1_R2 * (1 - bool_MG132)") |>
  addRC("R2 + TGFb", "R2_TGFb",
        "k_act_R2_TGFb * TGFb * R2 / (km_R2 + R2 + TGFb)",
        rateCompartment = "Cell") |>
  addRC("R2_TGFb", "R2 + TGFb", "k_deact_R2_TGFb * R2_TGFb") |>
  addRC("R2_TGFb", "R2_TGFb_int", "k_int_R2_TGFb * R2_TGFb") |>
  addRC("R2_TGFb_int", "TGFb", "k_decay_R2_TGFb_int * R2_TGFb_int") |>
  addRC("R2_TGFb_int", "", "k_dg_R2_TGFb_int * R2_TGFb_int * (1 - bool_MG132)") |>
  addRC("R1 + TGFb", "R1_TGFb",
        "k_act_R1_TGFb * TGFb * R1 / (km_R1 + R1 + TGFb)",
        rateCompartment = "Cell") |>
  addRC("R1_TGFb", "R1 + TGFb", "k_deact_R1_TGFb * R1_TGFb") |>
  addRC("R1_TGFb", "R1_TGFb_int", "k_int_R1_TGFb * R1_TGFb") |>
  addRC("R1_TGFb_int", "TGFb", "k_decay_R1_TGFb_int * R1_TGFb_int") |>
  addRC("R1_TGFb_int", "", "k_dg_R1_TGFb_int * R1_TGFb_int * (1 - bool_MG132)") |>
  addRC("R1 + R2_TGFb", "R1_R2_TGFb", "k_act_R1_R2_TGFb * R1 * R2_TGFb") |>
  addRC("R2 + R1_TGFb", "R1_R2_TGFb", "k_act_R2_R1_TGFb * R2 * R1_TGFb") |>
  addRC("R1_R2 + TGFb", "R1_R2_TGFb",
        "k_act_R1_R2_TGFb_direct * TGFb * R1_R2 / (km_R1_R2 + R1_R2 + TGFb)",
        rateCompartment = "Cell") |>
  addRC("R1_R2_TGFb", "",
        "(k_dg_R1_R2_TGFb + k_dg_R1_R2_TGFb_FB1 * FB2) * R1_R2_TGFb * (1 - bool_MG132)") |>
  addRC("Smad2", "pSmad2",
        "(k_phospho_pS2 / (1 + k_inh_pSmad2_FB2 * FB2)) * Smad2 * R1_R2") |>
  addRC("Smad2", "pSmad2",
        "(k_phospho_pS2 / (1 + k_inh_pSmad2_FB1 * FB2)) * Smad2 * R1_R2_TGFb") |>
  addRC("pSmad2", "Smad2", "k_dephos_S2 * pSmad2") |>
  addRC("Smad3", "pSmad3",
        "(k_phospho_pS3 / (1 + k_inh_pSmad3_FB2 * FB2)) * Smad3 * R1_R2") |>
  addRC("Smad3", "pSmad3",
        "(k_phospho_pS3 / (1 + k_inh_pSmad3_FB2 * FB2)) * Smad3 * R1_R2_TGFb") |>
  addRC("pSmad3", "Smad3", "k_dephos_S3 * pSmad3") |>
  addRC("pSmad2 + pSmad3 + Smad4", "C3",
        "k_form_S4Coip * pSmad2 * pSmad3 * Smad4") |>
  addRC("C3", "Smad2 + pSmad3 + Smad4", "k_dissolve_C3_dp2 * C3") |>
  addRC("C3", "pSmad2 + Smad3 + Smad4", "k_dissolve_C3_dp3 * C3")

reactions$compartments$Cell$volume      <- "1"
reactions$compartments$extraCell$volume <- "volumeEC"
reactions <- customTotals(reactions, list(totalSMAD2 = "Smad2 + pSmad2 + C3",
                                          totalSMAD3 = "Smad3 + pSmad3 + C3",
                                          totalSMAD4 = "Smad4 + C3"))
```

The readouts are the two receptor levels, the phospho- and total-SMAD levels, the
SMAD4 co-immunoprecipitation, the two receptor mRNAs and the ligand. In the
experiment they are measured on $\log_{10}$ scale, which has the same level sets as
$\mathrm{scale}\cdot h$, so they are supplied in that rational form. The experiment
toggles transcription (actinomycin D), translation (cycloheximide) and the
proteasome (MG132), or knocks a receptor down; each perturbation is one condition.

```{r smad-setup, eval = haveSympy}
observables <- eqnvec(
  R1_obs          = "scale_R1 * R1",
  R2_obs          = "scale_R2 * R2",
  pSmad2_obs      = "scale_pSmad2 * (pSmad2 + C3)",
  pSmad3_obs      = "scale_pSmad3 * (pSmad3 + C3)",
  TSmad2_obs      = "scale_TSmad2 * (Smad2 + pSmad2 + C3)",
  TSmad3_obs      = "scale_TSmad3 * (Smad3 + pSmad3 + C3)",
  Smad4_CoIP_obs  = "scale_CoIP * C3",
  TGFBR1_mRNA_obs = "scale_R1mRNA * R1mRNA",
  TGFBR2_mRNA_obs = "scale_R2mRNA * R2mRNA",
  TGFb_obs        = "scale_TGFb * TGFb")

events <- eventlist() |>
  addEvent(var = "TGFb",       time = 0,   value = "init_TGFb",      method = "replace") |>
  addEvent(var = "bool_CHX",   time = -30, value = "var_bool_CHX",   method = "replace") |>
  addEvent(var = "bool_MG132", time = -30, value = "var_bool_MG132", method = "replace") |>
  addEvent(var = "bool_ActD",  time = -30, value = "var_bool_ActD",  method = "replace")

# one condition per perturbation; each row bakes the switch values and the TGFb
# dose, and the receptor knockdowns rename a synthesis rate
cond.grid <- data.frame(
  Pertubation = c("Ctrl", "ActD", "CHX", "MG132", "R1Knd", "R2Knd"),
  init_TGFb = 1, stringsAsFactors = FALSE)
cond.grid$var_bool_ActD  <- ifelse(cond.grid$Pertubation == "ActD",  1, 0)
cond.grid$var_bool_CHX   <- ifelse(cond.grid$Pertubation == "CHX",   1, 0)
cond.grid$var_bool_MG132 <- ifelse(cond.grid$Pertubation == "MG132", 1, 0)
cond.grid$k_pr_R1mRNA <- ifelse(cond.grid$Pertubation == "R1Knd",
                                "k_pr_R1mRNA_R1Knd", "k_pr_R1mRNA")
cond.grid$k_pr_R2mRNA <- ifelse(cond.grid$Pertubation == "R2Knd",
                                "k_pr_R2mRNA_R2Knd", "k_pr_R2mRNA")
rownames(cond.grid) <- cond.grid$Pertubation
cond.grid$Pertubation <- NULL

# broadcast an identity trafo over the grid and insert each row's columns, giving
# one eqnvec per condition (a numeric bake for the switches/dose, a rename per knockdown)
cond.trafo <- eqnvec() |>
  define("x~x", x = getParameters(reactions, events)) |>
  branch(table = cond.grid, apply = "insert")
```

The boolean switches and the ligand are the externally driven inputs, declared
with `forcings`. The conditions enter as a per-condition `trafo` list, built by
broadcasting an identity map over the grid and inserting each row's covariates
with `branch()`; this is the general route when a perturbation renames a rate
(the receptor knockdowns) as well as fixing a switch. `equilibrate = TRUE` imposes
the pre-stimulus resting state with all forcings at zero and no events, solved
exactly over a finite field; the receptor-ligand complexes then fall out as zero
on their own, so nothing needs to be declared `fixed`, and the free Hill exponents
are recast to rational coordinates (Section \ref{sec-hill}) so the finite-field
solve still applies. The TGF$\beta$ dose at $t = 0$ is applied as an event on top
of that resting state, and the switch events at $t = -30$ open a second analytic
segment propagated exactly across the gap (Section \ref{sec-segments}); the
analysis starts at the earliest event time on its own. `cores` runs the
steady-state solves of the stacked conditions in parallel.

```{r smad-run, eval = haveSympy, warning = FALSE}
smad <- symmetryDetection(
  reactions, observables, method = "observability",
  events = events, trafo = cond.trafo,
  forcings = c("bool_ActD", "bool_CHX", "bool_MG132", "TGFb"),
  equilibrate = TRUE, reduceCQ = TRUE, reconstruct = TRUE, cores = 4)

summary(smad)
```

Six conditions, each split into two analytic segments by the pre-stimulus switch
events, are stacked, and the analysis reaches rank $65$ of $74$, leaving nine
non-identifiable directions. All nine are scalings, obtained exactly from the
stacked integer kernel, so each carries its integer weights, printed above; the
symbols in each weight are the factors it couples. Every free Hill exponent is
identifiable -- two survive only inside a weight. The nine fall into five groups,
each a product the data fix while its factors stay free.

* **Receptor mRNA synthesis against its readout gain** (two directions). For each
  receptor the mRNA synthesis rates (`k_pr_R1mRNA` and its knockdown variant
  `k_pr_R1mRNA_R1Knd`) and the protein synthesis rate `k_pr_R1` trade against the
  mRNA readout gain `scale_R1mRNA`; symmetrically for receptor 2.
* **Feedback protein against feedback mRNA** (three directions). Each feedback
  gene's protein synthesis `k_pr_FB*` trades against its mRNA synthesis
  `k_pr_FB*mRNA`, the unobserved intermediate mRNA absorbing the difference.
* **Linear feedback strength against feedback abundance** (one direction). `FB2`
  enters the phosphorylation and complex-degradation terms *linearly*
  (`k_inh_pSmad2_FB1`, `k_inh_pSmad2_FB2`, `k_inh_pSmad3_FB2`,
  `k_dg_R1_R2_TGFb_FB1`), so those action strengths confound with FB2's synthesis
  `k_pr_FB2` as an ordinary integer scaling.
* **Hill feedback strength, exponent-weighted** (two directions). The receptor
  mRNA feedbacks enter through `k_inh_R1mRNA_FB3 * FB3^nhill_R1`; the inhibition
  strength trades against the feedback synthesis `k_pr_FB3` with the free Hill
  exponent in the weight, $\xi_{k_{inh}} = -nhill_{R1}\,k_{inh}$ (and the same for
  receptor 2). The exponents `nhill_R1`, `nhill_R2` are themselves identifiable.
* **SMAD pool against the readout gains** (one direction). The conserved SMAD
  totals and the feedback Michaelis constants `Km_FB*mRNA` trade against the SMAD
  readout gains (`scale_pSmad2/3`, `scale_TSmad2/3`, `scale_CoIP`) and the
  complex-formation rate `k_form_S4Coip` (with weight $2$).

The boolean switches and the known TGF$\beta$ dose are constants of each condition
and appear in none of the directions. Each group is removed by the
reparametrisation shown earlier: fix a feedback gene's protein synthesis so the
mRNA synthesis absorbs it, provide an absolute receptor-mRNA measurement (or fix
`scale_R1mRNA`, `scale_R2mRNA`), and fix the readout gains against an absolute SMAD
calibration.


# Scope and limitations {#sec-scope}

* *Structural, not practical.* The methods use only $f$ and $g$. They
  cannot detect practical non-identifiability, which depends on the data
  and the noise.
* *Local rank versus explicit transformation.* `"observability"`
  decides local identifiability and names the unconstrained unknowns at
  generic points. `"polynomial"` returns the explicit finite transformation
  when it exists within the polynomial ansatz, at higher cost that grows
  with `pMax` and the number of variables.
* *Admissible observables.* `"polynomial"` requires a rational gradient of
  $g$, which includes rational maps and logarithms of rational maps.
  `"observability"` requires $f$ and $g$ themselves to be rational, since
  its finite-field construction is built from the field operations only; a
  logarithmic readout is supplied in the equivalent rational form
  $\mathrm{scale}\cdot h$ (Section \ref{sec-finitefield}), and a non-rational gradient
  (e.g. $\exp$, $\sqrt{\ }$) is outside the scope of both methods.
* *Conserved quantities.* With an `eqnlist` input and `reduceCQ = TRUE`,
  conserved moieties are eliminated before the search, which reduces the
  number of variables.
* *Reproducibility.* The observability point is chosen deterministically,
  so repeated runs give the same verdict; a custom evaluation point can be
  supplied through `point`.

# References

Ben-Or, M., and Tiwari, P. (1988). A deterministic algorithm for sparse
multivariate polynomial interpolation. In *Proceedings of the 20th Annual ACM
Symposium on Theory of Computing (STOC)*, 301--309.

Hong, H., Ovchinnikov, A., Pogudin, G., and Yap, C. (2020). Global
identifiability of differential models. *Communications on Pure and Applied
Mathematics* 73, 1831--1879.

Massonis, G., and Villaverde, A. F. (2020). Finding and breaking Lie symmetries:
implications for structural identifiability and observability in biological
modelling. *Symmetry* 12, 469.

Merkt, B., Timmer, J., and Kaschek, D. (2015). Higher-order Lie symmetries
in identifiability and predictability analysis of dynamic models.
*Physical Review E* 92, 012920.

Pohjanpalo, H. (1978). System identifiability based on the power series
expansion of the solution. *Mathematical Biosciences* 41, 21--33.

Saccomani, M. P., Audoly, S., and D'Angiò, L. (2003). Parameter identifiability
of nonlinear systems: the role of initial conditions. *Automatica* 39, 619--632.

Savageau, M. A., and Voit, E. O. (1987). Recasting nonlinear differential
equations as S-systems: a canonical nonlinear form. *Mathematical Biosciences*
87, 83--115.

Sedoglavic, A. (2002). A probabilistic algorithm to test local algebraic
observability in polynomial time. *Journal of Symbolic Computation* 33,
735--755.

Villaverde, A. F., Barreiro, A., and Papachristodoulou, A. (2016).
Structural identifiability of dynamic systems biology models. *PLoS
Computational Biology* 12, e1005153.
